How hibernation in frogs drives brain and reproductive evolution in opposite directions

Version 1

Abstract

Abstract

Environmental seasonality can promote the evolution of larger brains through cognitive and behavioral flexibility but also hamper it when temporary food shortage is buffered by stored energy. Multiple hypotheses linking brain evolution to resource acquisition and allocation have been proposed, albeit separately for different groups of birds or mammals rather than being directly compared within any single group. Here, using direct tissue measurements and experimentally validated brumation (‘hibernation’) parameters, we integrated these hypotheses across frogs in the context of varying brumation duration and its environmental correlates. We show that protracted brumation reduces brain size and instead promotes reproductive investments, likely in response to brumation-dependent changes in the socio-ecological context that ultimately affect the operation of sexual selection and evolution of mating systems. Our results reveal novel insight into the complex processes of brain and reproductive evolution in organisms whose ‘cold-blooded’ metabolism is particularly susceptible to environmental seasonality.

Introduction

Seasonal fluctuations in climatic conditions and primary productivity can result in temporary food limitation, imposing energetic constraints on animals. Maintaining a positive energy balance across seasons, or at least minimizing the negative balance during lean periods, can be achieved by more constant net energy intake than predicted solely by food abundance (Sol, 2009), or by reduced investments in costly organs (Heldstab et al., 2018). At the center of both strategies is the size of the brain. On the one hand, a relatively large brain can improve the cognitive ability and behavioral flexibility (Reader and Laland, 2002; Lefebvre et al., 2004; Sol et al., 2005; Benson-Amram et al., 2016) to locate more diverse and dispersed food sources to buffer the environmental fluctuations of seasonal habitats (‘cognitive buffer hypothesis’) (Allman et al., 1993; Sol, 2009; van Woerden et al., 2012). On the other hand, as the high metabolic costs of brain tissue (Mink et al., 1981; Aiello and Wheeler, 1995; Lukas and Campbell, 2000) may not be temporarily reducible (Mink et al., 1981), periodic food scarcity is expected to constrain brain size evolution (‘expensive brain hypothesis’) (Isler and van Schaik, 2009).

Extreme fluctuations in resource acquisition and total metabolic activity are found in hibernating species, whose basal metabolic rate may drop by over 90% during hibernation (Ruf and Geiser, 2015). Such a radical reduction in metabolism is likely to limit investments in the maintenance of brain function and thus the support of a large brain. Indeed, mammals with some period of hibernation tend to have relatively smaller brains than their non-hibernating counterparts (Heldstab et al., 2018).

In addition to cognitive responses (van Woerden et al., 2010), periods of food scarcity can also promote the evolution of physiological responses. For example, a longer digestive tract may permit more efficient resource accumulation during a short active period (Sibly, 1981) and so could be favored by selection in species with prolonged hibernation. This response could parallel brain size evolution or result in an evolutionary trade-off between the two organs (‘expensive tissue hypothesis’) (Aiello and Wheeler, 1995), similar to predicted trade-offs between brain size and sexually selected traits (‘expensive sexual tissue hypothesis’) (Pitnick et al., 2006) or other costly organs more generally (‘energy trade-off hypothesis’) (Isler and van Schaik, 2006). Further, physiological buffering is often accompanied by a seasonal reduction in the metabolic rate or activity (e.g., hibernation), with energy drawn from stored fat reserves (Heldstab et al., 2016). Even though hibernation does not preclude benefits of cognitive abilities (e.g., to efficiently accumulate fat reserves before hibernation), the evolution of a relatively larger brain could be hampered by seasonally alternating between cognitive benefits during the active period and maintenance costs at no obvious benefit during hibernation. In fact, it has been proposed that the benefits of maximal fat stores and minimal metabolic expenditure on somatic maintenance during hibernation, on average, are likely to outweigh those of a large brain (‘fat−brain trade-off hypothesis’) (Navarrete et al., 2011; Heldstab et al., 2016). By comparison, non-hibernators may always gain a positive net benefit of a relatively large brain, possibly even enhanced if it mitigates resource acquisition when food is seasonally scarce (Allman et al., 1993; Sol, 2009; van Woerden et al., 2012). If so, the probability of positive selection on brain size should be higher in the absence of hibernation, providing one explanation for the relatively smaller brains in hibernating compared to non-hibernating mammals (Heldstab et al., 2018).

Allocating a fixed energy budget to the demands of different organs throughout prolonged hibernation might also have important reproductive consequences. Particularly for species that reproduce almost immediately after emerging from hibernation, as in some mammals (Psenner, 1957; Place et al., 2002) and many amphibians (Wells, 1977; Fei and Ye, 2001), the seasonal recrudescence of their reproductive tissue necessarily occurs before emergence when the stored resources are most limited (Isler and van Schaik, 2009; Isler, 2011). Reproductive investments, however, are intimately linked to fitness, such as testis size that may be under intense selection by sperm competition resulting from female multiple mating (Lüpold et al., 2020). Female promiscuity itself is prevalent where males are less able to effectively monopolize their mates (Lüpold et al., 2014), and this would seem particularly likely when the breeding activity is highly synchronized in dense aggregations (Lüpold et al., 2014, 2017). Indeed, across anurans (frogs and toads) that are often bound to small water bodies for reproduction, males invest relatively more in their testes and less in their forelimbs (used in pre-mating competition) as population density increases (Buzatto et al., 2015; Lüpold et al., 2017). If the breeding activity is more synchronized because of a shortened active period, thus increasing the risk of either competitive fertilization or simply of sperm depletion by a high mating rate (Vahed and Parker, 2012), selection for relatively larger testes would be stronger precisely where the available fat stores need to last longer, with likely consequences for resource demands and allocation while overwintering. Including reproductive investments and breeding patterns in studies of allocation trade-offs in response to hibernation and environmental seasonality would thus seem critical but remains to be done, particularly in the context of brain evolution.

The opposing selection pressures on brain or gonad size (i.e., cognitive or fitness benefits versus metabolic costs), varying degrees of seasonality and diverse strategies of buffering periodic food scarcity (e.g., cognitive versus physiological) between species render environmental fluctuations an ideal context to study brain size evolution. The different hypotheses invoked to explain the coevolution of brain size with other organs were each independently developed for a separate set of mammalian or avian taxa, the two vertebrate classes with the relatively largest brains (Jerison, 1973). These hypotheses have yet to be directly tested against one another in a single taxon and ideally in the immediate context of seasonal activity, considering the extent, rather than the mere presence/absence, of hibernation. This last point is important to the extent that the classification of ‘hibernation’ could range anywhere between one or multiple brief inactive bouts (i.e., minimally different from non-hibernating species) and spending most months of the year in dormancy, with severe energetic and life-history constraints despite again simply being classified as hibernating. Further, understanding the generality of the patterns reported in the large-brained mammals or birds requires validation in other taxa, ideally with smaller brains and different overall energetic demands. Such generalization would also permit contextualizing brain evolution in the two largest-brained taxa in terms of the strength of selection on encephalization relative to potential metabolic constraints.

A particularly suitable system is presented in ectothermic (‘cold-blooded’) species whose metabolism and activity are tightly linked to their ambient temperature and considerably hampered outside a species-specific temperature range (Wells, 2007). Hence, temperature constraints on metabolism and activity patterns potentially set stricter physical boundaries to the ability to buffer food scarcity through behavioral flexibility than in the endothermic (‘warm-blooded’) mammals or birds. Additionally, the different groups of mammals, for which most hypotheses on brain evolution were proposed, exhibit extremely diverse Bauplans and lifestyles that could confound overall conclusions. By contrast, anurans are relatively homogeneous in body size and shape, diet or locomotion (Kardong, 2019), but still markedly divergent in their relative brain size (Liao et al., 2022) or in reproductive (Lüpold et al., 2017) and other investments (Wells, 2007) in response to environmental variation. Consequently, any resource trade-offs around the evolution of brain size could, if they exist, be easier to isolate across anurans than across mammals.

Here, we examine variation in male brain and testis size relative to body fat, limb muscles and the main visceral organs (see Fig. 1) across the males of 116 anuran species in the context of the vastly varying ‘hibernation’ periods and their environmental correlates. Frogs differ from mammals in their physiology of hibernation, in that the seasonal inactivity is a consequence of ambient temperatures dropping below the activity range rather than of active metabolic depression (referred to here as ‘brumation’ for distinction) (Pinder et al., 1992; Wells, 2007). Yet, important parallels remain in terms of resource allocation. In both taxa, hibernating animals spend extended periods depleting fixed, previously accumulated energy stores across sequential investments, albeit at a reduced metabolic rate (Staples, 2016). By contrast, non-hibernators can at least partially compensate for spent resources as they go, but their higher metabolic rate requires continued resource acquisition even when food is seasonally scarce (Heldstab et al., 2016). Such differences between strategies are likely to affect the relative costs and benefits of different organs, and thus how species optimally allocate resources between them. Relatively larger traits (e.g., brain) can evolve—possibly at the cost of other traits—if they confer some net fitness benefits.

Allometric and seasonal variation in the mass of 11 organs and tissues. A: Allometric slopes between the mass of each tissue and cubed snout-vent length (SVL3) so that proportionate scaling follows a slope of 1. Each point represents a species-specific mean value in breeding condition (N = 116). Relationships deviating from proportionate scaling (based on bootstrapped 95% confidence intervals) are highlighted in blue (steeper than unity) or red (shallower than unity). B: Mean percent change with 95% confidence interval for body mass and each individual tissue of 50 anuran species with data from both shortly before and after brumation (= breeding), based on absolute tissue masses between stages and log-transformed to maintain symmetry and additivity (Törnqvist et al. 1985): log(post-brumation / pre-brumation) × 100. The transparent grey dots depict the species-specific values. C: The effect of the brumation period on the percent mass change for the two tissues with a substantial seasonal variation (see panel b): body fat (top) and testes (bottom). Each point indicates a species.

Brumating anurans drop their heart rate, become sluggish, draw the nictitating membranes across the eyes for protection, spread their legs for stability, and undergo multiple physiological changes to protect against freezing or to switch from pulmonary to cutaneous gas exchange or from aerobic to anaerobic metabolism (Pinder et al., 1992; Fei and Ye, 2001; Wells, 2007; Tattersall and Ultsch, 2008). Yet, several anuran species are known to regularly move in their burrows or underwater hibernacula in response to changes in soil temperature or oxygen concentration (van Gelder et al., 1986; Stinner et al., 1994; Holenweg and Reyer, 2000), or when disturbed (Tattersall and Ultsch, 2008; Niu et al., 2022). Brumating frogs can also exhibit higher levels of brain cell renewal than active ones, possibly to avoid brain damage (Cerri et al., 2009), and it seems plausible that the total investment in such renewal would increase with the amount of brain tissue. Further, larger brains could also be less tolerant to the often hypoxic brumation conditions (Pinder et al., 1992; Wells, 2007; Tattersall and Ultsch, 2008) if the findings from other ectotherms extend to brumating anurans (Sukhum et al., 2016). Overall, maintaining a relatively larger brain while brumating is likely to come with greater costs that could constrain brain evolution compared to species that show only brief or no seasonal inactivity.

Instead of the typical indirect proxies or scores (Navarrete et al., 2011; Heldstab et al., 2016, 2018; Luo et al., 2017), we directly quantified seasonal changes in tissue sizes and estimated the brumation duration as the period during which temperatures were continuously below the species-specific activity threshold. These estimated brumation durations corresponded to the periods of no frog activity detected in field surveys. Our results provide robust evidence that the duration of brumation under varying climates modifies how frogs allocate their limited resources. We then resolve the direct and indirect links between these variables to test the different hypotheses on brain evolution, which have been independently proposed for different avian and mammalian taxa, within our single set of anurans. We further integrate variation in breeding contexts and reproductive evolution in this context of resource allocation. Our broad comparative approach permits disentangling important evolutionary responses to environmental seasonality with its far-reaching behavioral and physiological consequences for species during both their active and inactive periods.

Results

Determinants of brumation duration.

For a mean (±SD) of 3.41 ± 0.95 males in each of 116 anuran species, we combined experimentally determined thermal activity thresholds with multi-year temperature fluctuations at their collection sites to estimate the species-specific brumation periods (details and validation in Material and Methods). These periods averaged between 0.6 ± 0.5 and 250.5 ± 16.7 days across the five years examined, with high repeatability within species (R = 0.95 [95%CI: 0.93, 0.96]; Fig. S1; Table S1). Twenty-two species were predicted to overwinter for ≤9 days, another three species for 18−27 days, and the remaining 91 species for ≥47 days, resulting in a rapid shift between 9 and 47 days. Since the ground microclimate may buffer some of the fluctuations in air temperature and frogs can endure short cold spells without dormancy, we conservatively considered the 25 species with ≤27 days below their experimental temperature threshold as unlikely to show any sustained brumation.

In phylogenetic regressions (Freckleton et al. 2002), the brumation period increased with both latitude and elevation of the study sites, as well as with the variation in temperature (r ≥ 0.42, t114 ≥ 5.00, P < 0.001; Table S2). By contrast, the brumation period was inversely related with the annual mean temperature and precipitation, and the duration of the dry season (r < −0.29, t114 < −3.24, P < 0.002), but not significantly associated with longitude or the variation in precipitation (r < 0.09, t114 < 0.95, P > 0.34; Table S2). Except for the period of the dry season, all these results were qualitatively identical when focusing only on the 91 species with some expected brumation period (see above; Table S3). Among these 91 species, those inhabiting cooler and more seasonal climates entered and emerged from their inactive state at lower temperatures (Table S4), suggesting an increased cold tolerance to maximize their active period.

Effect of brumation on individual tissue investments

Based on this variation in the periods of active resource acquisition or metabolizing stored resources, respectively, we explored potential consequences for resource allocation in the same set of males across the 116 species. In phylogenetic regressions (Freckleton et al. 2002), neither snout-vent length (SVL) nor body mass covaried with the duration of brumation or any other environmental variable (|r| ≤ 0.15, |t114| ≤ 4.03, P ≥ 0.11; Table S5), with only a weak, non-statistically significant trend toward reduced body mass at higher elevation (r = −0.18, t114 = −1.91, P = 0.06, phylogenetic scaling parameter λ = 0.94 [95%CI: 0.85, 1.00]). The different organs and tissues themselves all increased with body size, but with different allometric slopes. Whereas the mass of body fat and of both the forelimb and hindlimb muscles showed a disproportionately steep increase relative to body size across the same 116 species (all β ≥ 1.10 [1.03, 1.18]), the allometric slope was substantially shallower for brain size (β = 0.49 [0.44, 0.54]; Table S6; Fig. 1A). All remaining tissues did not deviate from proportionate scaling (i.e., 95%CI including 1.00; Table S6; Fig. 1A). Hence, the evolution of brain size appears to be more constrained than that of other organs when selection favors larger body size.

With these different investments in brain size compared to other expensive tissues, we next tested if the evolution of brain size is constrained by extended brumation, similar to what has been suggested for mammals based on the mere presence or absence of any hibernation (Heldstab et al. 2018). Absolute brain size was independent of the brumation period (r = −0.10, t115 = −1.08, P = 0.28, λ = 0.89 [0.73, 0.97]), but controlling for SVL as a proxy of body size, males of species with protracted hibernation had relatively smaller brains, whether quantified during the breeding season (partial r, rp = −0.31, t113 = −3.50, P < 0.001, λ = 0.35 [0.00, 0.61]; Fig. 2A, Table S7) or shortly before entering hibernation (N = 50 species means based on 2.64 ± 0.94 males each: rp = −0.51, t47 = −4.03, P < 0.001, λ = 0.00 [0.00, 0.56]; Table S7). This trend was not a mere result of changing body size in response to brumation, as the SVL itself was independent of the hibernation period (see above; Table S5). On average, brumating species tended to have relatively smaller brains than those that are unlikely to overwinter for an extended period (Table S8), similar to the study of mammals that used the presence/absence of hibernation as a predictor (Heldstab et al. 2018). However, the above trend also applied to those 91 species that we classified as brumating for some period (Table S9), thus providing stronger evidence for a link between brumation and brain evolution than the coarse binary classification. This pattern further remained robust when we recalculated the predicted brumation periods such that frogs were only considered to enter their inactive state at temperatures that were 2°C or 4°C below their experimentally derived thresholds. These two more conservative thresholds for brumation simulated a potential buffering effect of underground burrows and other shelters relative to the seasonal variation in air temperatures that was available in meteorological databases (details and validation in Material and Methods; Table S10).

Effects of brumation duration on the relative tissue sizes. Relationships between brumation duration and the relative mass of the brain (A), body fat (B), and testes (C) across males of 116 anuran species in breeding (post-brumation) condition. All axes are controlled for the snout-vent length and phylogeny.

Species with a prolonged brumation period further had relatively more body fat (rp ≥ 0.25, t113 ≥ 2.72, P ≤ 0.008; Fig. 2B) and, at least in breeding condition, relatively larger testes (rp = 0.36, t113 = 4.06, P < 0.001, λ = 0.77 [0.40, 0.90]; Fig. 2C) and relatively smaller hindleg muscles (rp = −0.22, t113 = −2.37, P = 0.02, λ = 0.22 [0.00, 0.51]; Tables S7). These patterns were generally consistent when using the presence/absence of brumation (Table S8), buffered temperature fluctuations (Tables S10), or excluding the 25 species that are unlikely to overwinter (except for the non-significant effect on body fat; Table S9). The size of the remaining tissues was independent of brumation (Tables S7 to S10) and did not change significantly between the two sampling periods (Fig. 2).

Comparisons between pre- and post-brumation males, on average across all species, revealed a ca. 50% decline in fat tissue and 100% increase in testis size, indicating resource depletion and testicular recrudescence while brumating, respectively (Fig. 1B). Of the remaining expensive tissues, only the 95%CI of brain size excluded zero, but this change was small compared to the two traits above and within the range of the many tissues with no significant change. Thus, the biological significance of this putative increase in brain size during brumation is questionable as it could simply be attributable to general differences between the few individuals per species that were sampled during each period. We will thus refrain from further interpretation. Among the two tissues with considerable change, the extent of fat depletion increased significantly with the brumation period (r = −0.36, t48 = −2.68, P = 0.01, λ = 0.00 [0.00, 0.55]), as did that of testis regrowth (r = 0.39, t48 = 2.89, P = 0.006, λ = 0.82 [0.16, 0.98]; Fig. 1C).

Since the patterns for relative testis size might be a response to a shorter, more synchronized mating season when brumation is long (Wells 2007), we tested for links between brumation duration and different breeding parameters. Here, prolonged brumation shortened the breeding season (t41 = −4.47, P < 0.001, λ = 0.00 [0.00, 0.38]; Fig. S2A), with a much stronger effect than any of the climatic variables (Table S11), particularly when considered jointly (Table S12). Hence, the effect of these climatic variables may be indirect, mediated by brumation. A shorter breeding season further increased the probability of dense breeding aggregations (phylogenetic logistic regression: N = 42, z = −3.03, P = 0.002, α = 0.02; Fig. S2B). That brumation might mediate the effects of climatic variables on breeding aggregations via the duration of the breeding season was also supported by a phylogenetic confirmatory path analysis (von Hardenberg and Gonzalez-Voyer 2013; Gonzalez-Voyer and von Hardenberg 2014) (Fig. S3; Table S13). Finally, when combining these data with previously published data on the density of breeding populations (Lüpold et al. 2017) (N = 8 species overlapping), this effect was also supported by a trend toward higher mean population densities in species with a shorter breeding season (r = −0.69, t6 = −2.37, P = 0.06, λ = 0.00 [0.00, 1.00]), albeit based on a small sample size (Fig. S2C).

To explore possible causal links between these breeding parameters and variation in relative testis size, we conducted directional tests of trait evolution (Pagel 1994; Revell 2012), which test if changes in two traits are unilaterally dependent, mutually dependent, or independent (Pagel 1994). Although independent evolution was the best-supported scenario based on the Akaike Information Criterion (AIC), the model with changes in relative testis size dependent on those in the breeding season was not significantly different (ΔAIC = 0.83, wAIC = 0.33 compared to independent model with wAIC = 0.50) and suggested that testis size was most likely to increase in response to a shortened breeding season (Fig. S4A). A model predicting increases in relative testis size in response to aggregation formation pointed in a similar direction (Fig. S4B), albeit not below a ΔAIC cut-off of 2 (ΔAIC= 2.38, wAIC = 0.20 compared to independent model with wAIC = 0.66). Hence, it is at least possible that breeding conditions could mediate the positive relationship between the brumation period and relative testis size in our relatively small sample of species.

Covariation between different tissues

Since all tissues depend on the same finite resources, their evolution in response to brumation is unlikely to be independent. In pairwise partial correlations controlling for SVL and phylogeny, all tissue masses covaried either positively or were not significantly associated with one another (Fig. S5). As such, our data do not support the expensive tissue (Aiello and Wheeler 1995) or the more general energy trade-off hypotheses (Isler and van Schaik 2006), which predict trade-offs of brain size with the size of the digestive tract or other costly organs, respectively. However, considering that brain size differed from fat tissue, hindlimb muscles and testes in the direction of allometric relationships or their responses to brumation duration, respectively, pairwise correlations may not reveal more complex allocation patterns. To examine the relative investments in these four most informative tissues simultaneously, we partitioned the total body mass into the proportional representation of each of these tissues (and the remaining mass combined as a control) to generate a five-variable compositional dataset (van den Boogaart and Tolosana-Delgado 2013). The combined mass of all four focal tissues scaled proportionately with body size (allometric β = 1.02 [0.96, 1.07], λ = 0.01 [0.00, 0.11]), confirming that each species allocated a size-independent share of its total resources to the four focal tissues combined. How these investments were distributed across these tissues, however, varied considerably between species. In pairwise correlations between the four focal tissues, transformed to centered log ratios (van den Boogaart and Tolosana-Delgado 2008) and controlling for phylogeny, brain mass covaried negatively with both body fat and testis mass, whilst testis mass was negatively correlated with hindlimb muscle mass but not associated with body fat (Fig. 3A-D; Table S14). To further examine the effect of brumation duration on all five variables simultaneously, we conducted a phylogenetic multivariate regression analysis (Clavel et al. 2015) on the same compositional data, but now transformed to isometric log ratios as recommended for multivariate models (van den Boogaart and Tolosana-Delgado 2008, 2013). Brumation duration had a significant effect on the body composition of frogs (Pillai’s trace = 0.33, effect size ξ2 = 0.30, P = 0.001). The back-transformed coefficients of body fat (0.23) and testis mass (0.26) were greater, and those of brain mass, hindlimb muscles and the rest of the body were lower (0.16, 0.17, and 0.18, respectively), than the expected coefficient of 0.20 for all five variables if increasing brumation duration were to cause no compositional change (also see Fig. 3).

Effects of brumation duration on the relative tissue sizes. Panels A-D depict the phylogenetic correlations (shown as phylomorphospace plots (Revell 2012)) between the relative masses of (A) brain and body fat, (B) brain and testes, (C) testes and body fat, and (D) testes and hindlimb muscles, respectively, across the 116 species (results in Table S14). The relative tissue masses represent the centered log ratios of the compositional data, and the lines connect the nodes of the underlying phylogeny, indicating that phenotypic correlations are not simply the result of phylogenetic clustering. The correlation coefficients and 95% confidence intervals are indicated. The loadings from a phylogenetic principal component analysis (Revell 2012) on the same variables are also mapped as vectors onto biplots between (E) the first and second or (F) the second and third principal components. In all panels, the point colors reflect the species-specific brumation periods (see legend in panel A). Generally, where brumation was relatively shorter or absent, species also tended to have relatively larger brains, less body fat and smaller testes, respectively, consistent with the univariate analyses (Fig. 2).

We further confirmed the negative associations of brain size with both body fat and testis mass, and that between testis mass and hindlimb muscles, in a phylogenetically informed principal component analysis (Revell 2012) (Fig. 3E,F, Table S15). Here, the first three principal components (PC1 to PC3) explained 84.7%, 8.0% and 5.1% of the total variance, respectively. Even though both PC2 and PC3 explained a relatively small proportion of the total variance, they separated the different tissues. PC2 was predominantly loaded by brain size (0.30) and testis mass (−0.44), and PC3 by brain size (0.26) and body fat (−0.31). Both cases thus indicated a negative association between the pairs of traits within the multivariate trait space (Fig. 3F). Further, brumation duration covaried negatively with PC2 (r = −0.53, t114 = −6.59, P < 0.0001, λ = 0.74 [0.49, 0.92]), consistent with a decrease in brain size and increase in testis mass towards longer brumation, but it was not significantly associated with PC3 (r = −0.12, t114 = −1.30, P = 0.20, λ = 0.41 [0.00, 0.70]).

Direct and indirect effects revealed by path analysis

To disentangle the evolutionary links between the relative sizes of costly tissues, and to test the most prominent hypotheses of brain evolution simultaneously in the direct context of brumation, we finally integrated these patterns in a phylogenetic confirmatory path analysis (von Hardenberg and Gonzalez-Voyer 2013; Gonzalez-Voyer and von Hardenberg 2014) based on 28 pre-specified candidate path models (Fig. S6; Table S16). The averaged model (Fig. 4) confirmed the negative effect of prolonged brumation on relative brain size (β = −0.15 [−0.22, −0.07]), which paralleled direct (β = 0.16 [0.07, 0.26]) and indirect positive effects on relative testis size (Fig. S7). These indirect effects on testis evolution were mediated by the relative amount of adipose tissue, which increased with both the brumation period (β = 0.13 [0.06, 0.20]) and the relative size of the digestive tract (β = 0.51 [0.37, 0.65]), and in turn had a positive effect on relative testis size (β = 0.39 [0.18, 0.59]).

Results of the averaged phylogenetic path model. Visual representation of the average phylogenetic path model across 116 anuran species. Arrows reflect the direction of the path, with their widths being proportional to the standardized regression coefficients and colors indicating the sign (blue = positive, red = negative). Paths with 95% confidence intervals excluding 0 (i.e., arrows highly probable) are drawn as solid arrows, all others as dashed, semi-transparent arrows. For simplicity and to avoid over-parameterization, other organs were omitted in path models as they showed little covariation with brumation duration or brain size. All phenotypic traits were log-transformed, and all variables were controlled for body size via additional paths from log SVL. Although SVL had a strong effect on all variables (all β > 0.37), its thick blue arrows to each box are omitted in this figure only for visual clarity, but all path coefficients are presented with their 95%CI in Fig. S7, with further details in Fig. S6 and Table S16.

Discussion

Our experimentally validated brumation periods and direct measures of the relative sizes of, and changes in, expensive tissues revealed novel insight into the complex and non-independent processes of brain and reproductive evolution in anurans whose ‘cold-blooded’ metabolism is particularly susceptible to environmental seasonality. Species in highly seasonal environments, which go through prolonged inactive periods, had relatively smaller brains than those in more stable climates. By reducing the brain tissue and its associated maintenance costs, brumating species redirected their additional fat reserves to reproduction, possibly due to the shorter breeding season with its socio-ecological consequences.

We demonstrated that anurans inhabiting cooler and more seasonal climates entered and emerged from their inactive state at lower temperatures, indicating increased cold tolerance to maximize their active period. Yet, with the metabolic rate depending on its thermal environment in ectotherms, low temperatures might render foraging and digestion too inefficient to extend activity beyond certain thresholds (e.g., Riddle, 1909; Fontaine et al., 2018). In response to the varying periods of a positive versus negative net energy balance, we found that species with protracted brumation have relatively smaller brains. These results confirm that, unlike birds (Sol, 2009) and some mammals (van Woerden et al., 2010), challenging and unpredictable environmental conditions select for physiological rather than cognitive buffering in anurans (Luo et al., 2017). Supporting a relatively large brain may not be sustainable in the absence of continued resource intake, or large brains could be less tolerant to hypoxic conditions (Sukhum et al., 2016) during brumation. However, selection for relatively larger brains may also simply be stronger in species with a long active (and so short brumation) period owing to extended cognitive benefits such as predator evasion (Kotrschal et al., 2015) or exploitation of better and more diverse food sources (Lefebvre et al., 1997; Jiang et al., 2022).

In pairwise comparisons, the relative sizes of the tissues examined here, including the brain, were generally positively correlated. These results reject both the expensive tissue (Aiello and Wheeler, 1995) and the more general energy trade-off hypotheses (Isler and van Schaik, 2006), which predict trade-offs of brain size with the size of the digestive tract or other costly organs, respectively. This lack of support in anurans aligns with a previous report in mammals (Navarrete et al., 2011) despite their substantially smaller brains and vastly different ecology and physiology, including a lower metabolic rate and largely lacking physiological thermoregulation. When focusing jointly on the four tissues (brain, body fat, testes, hindlimb muscles) that covaried with brumation duration, however, relative brain size covaried negatively with the relative mass of both the fat tissue and testes as predicted by the fat−brain trade-off (Navarrete et al., 2011) or expensive sexual tissue hypotheses (Pitnick et al., 2006), respectively. Species with an extended brumation period exhibited both a relatively larger amount of body fat in total and a higher degree of its depletion, providing direct evidence to the hypothesis that anurans buffer lean periods by metabolizing stored fat (Luo et al. 2017; Huang et al., 2020). Although adipose tissue may not itself be metabolically expensive, transporting it adds costs to locomotion, particularly when jumping away from predators (Moreno-Rueda et al., 2020) or climbing trees compared to moving horizontally on land or in water (Alexander, 2003; Hanna et al., 2008). Consistent with this notion, arboreal species tended to be leaner compared to (semi)aquatic or terrestrial species (Table S17), controlling for brumation duration and relative brain size, both of which we had shown to covary with body fat (Figs. 2 and 3).

In addition to a relatively smaller brain and relatively larger fat reserves, species with a prolonged brumation period also had relatively smaller hindleg muscles and larger testes. Since anurans move primarily using their hindlegs, the negative relationship between hindleg muscle mass and brumation duration may be linked to more movement during a longer active period, including predator evasion (Marchisin and Anderson, 1978; Liao et al., 2022). Further, one explanation for the negative association between brumation and relatively testis size could be that a shortened active period compacts the breeding season, resulting in denser and more aggregated breeding populations and likely more synchronous mating activity (Wells, 2007), as indicated by our path analysis. These aggregations heighten the likelihood that multiple males attempt to mount the same female simultaneously (Lüpold et al., 2017 p. 20), resulting in more intense male-male competition over fertilization and thus enhanced investments in sperm production (Liao et al., 2018; Lüpold et al., 2020). Our results thus reveal how the environmental variation and physical constraints that determine the species-specific brumation pattern might play a pivotal, albeit previously overlooked, role in shaping the socio-ecological context of breeding, the mode and degree of sexual selection, and ultimately the evolution of mating systems, broadening Emlen & Oring’s (Emlen and Oring, 1977) general predictions.

Species with protracted brumation not only exhibited relative larger testes, but also a greater change in testis size from pre-to post-brumation (i.e., breeding) condition. The testes of seasonally breeding anurans regress after the mating season and regrow before the next (Ogielska and Bartmańska, 2009). Whereas non-brumating species can compensate for the resources invested in testicular recrudescence by energy uptake, those with a short breeding season following a prolonged inactive period depend on the stored fat to regrow their testes before or immediately after emergence from their hibernaculum. Hence, resources are diverted away from the brain and other organs, which may be the case especially in species such as Brachytarsophrys spp., in which the fully developed testes combined weigh 12−14 times more than the brain (Data S1).

To integrate all these different patterns and test the most prominent hypotheses of brain evolution simultaneously in the direct context of brumation, we also performed a phylogenetic path analysis. This analysis corroborated the negative effect of brumation duration on relative brain size and revealed both its direct and indirect positive effects on relative testis size. The indirect effects on testis evolution were mediated by the amount of adipose tissue, which itself responded to variation in the inactive period (energetic demand) and the size of the digestive tract (energy uptake). That variation in body fat did not contribute to brain size evolution in this more comprehensive analysis compared to pairwise correlations suggests that the fat−brain trade-off may not be a direct one. Rather, prolonged brumation, and thus short active period, may enhance selection on fat storage for testicular investments in addition to starvation avoidance, while simultaneously selecting for smaller brains (or weakening selection for larger brains) due to a shifted balance between the cognition-derived fitness benefits and the energetic costs related to brain size (Fig. 3).

A trade-off between brain and testis sizes has been reported for bats (Pitnick et al., 2006), albeit unsupported by a later study (Dechmann and Safi, 2009) or in other mammalian groups (Lemaître et al., 2009). In anurans, the apparent brain−testes trade-off may not be a direct functional one but result indirectly from opposing selection on brain and testis sizes via environmental seasonality and relative durations of the active and inactive periods. The testes may primarily evolve in response to the heightened levels of sperm competition and sperm depletion during the shortened and more synchronized breeding season. The brain, while also responding to sexual selection (Mai et al., 2020), is central to various activities other than mating, including feeding (Lefebvre et al., 1997) or predator avoidance (Kotrschal et al., 2015; Liao et al., 2022) that are themselves subject to climatic conditions and may independently influence brain evolution. In addition, whereas testes can regress to save energy when inactive (Ogielska and Bartmańska, 2009), brain metabolism may be less reducible (Mink et al., 1981), resulting in a different balance between fitness costs and benefits between these organs in relation to seasonality.

In conclusion, our analyses resolve how brumation in anurans, resulting from high environmental seasonality, may constrain the evolution of brain size and affect, directly or through its environmental correlates, resource allocation between costly tissues. These results reveal novel insight into the complex context of brain size evolution in far smaller-brained organisms than those typically studied, and whose ‘cold-blooded’ metabolism is particularly susceptible to environmental fluctuations. Our data also draw attention to the impact that varying brumation periods are likely to have on the operation of sexual selection and mating system evolution by modifying the timing and socio-ecological context of breeding during the active period. In turn, these factors determine reproductive investments and, via differential resource allocation, may also affect brain evolution. These non-independent selective processes promoting diversification in different traits highlight the need to study the evolutionary trajectory of a given trait such as brain size in the immediate context of both simultaneous investments to other tissues and the species-specific ecology.

Materials and Methods

Sample collection and preparation

Between 2010 and 2020 and as part of concurrent studies, we collected a total of 396 sexually mature males from 116 anuran species (3.41 ± 0.95 males each) in post-brumation breeding condition and an additional 132 adult males from 50 of these species (2.64 ± 0.94 males each) shortly before entering their hibernacula (Data S1 and S2). For each species, we sampled all males at a single location in southern and western China with known longitude, latitude, and elevation (Data S3). Upon transfer to the laboratory, we sacrificed the individuals by single-pithing, measured their snout-vent length (SVL) to the nearest 0.01 mm with calipers and then preserved them in 4% phosphate-buffered formalin for tissue fixation.

After two months of preservation, we weighed each complete specimen to the nearest 0.1 mg using an electronic balance to obtain body mass before dissecting them following a strict protocol. We separately extracted the brain, heart, liver, lungs, kidneys, spleen, digestive tract, testes, limb muscles, and fat stores, cleaned these tissues and immediately weighed them to the nearest 0.1 mg with an electronic balance. We additionally measured the length of the digestive tract to the nearest 0.01 mm using calipers. We excluded emaciated individuals or those exhibiting visible organ pathologies from our analyses.

Environmental seasonality

For each collection site, we retrieved from the 30-year climate history of https://www.meteoblue.com the monthly mean temperature (in °C) and total precipitation (in mm) (Data S3) and used these values to calculate location-specific annual means and coefficients of variation. We also determined the duration of the dry season, P2T, as the number of months, for which the total precipitation was less than twice the mean temperature (Walter, 1971).

Brumation period

One way that anurans can physiologically respond to seasonality is by adjusting their thermal sensitivity and thus brumation period (Wells, 2007), which in turn could directly or indirectly affect the evolution of brain size (Heldstab et al. 2018). Hence, we estimated the brumation period for all 116 species. To this end, we visited the field sites for 30 of our species daily around the expected start and end times of brumation (based on prior experience). For each species, we recorded the dates and temperatures (using a Kobold HND-T105 high-precision thermometer to the nearest 0.1°C) when the last frogs of a given species were seen at the end of their active period (with no further activity detected for at least seven days) and when the first individuals were detected in the spring. For the same 30 species (and using the same individuals as for morphological measurements), we then experimentally simulated brumation using a Q18 temperature-controlled refrigerator in Shenzhen Pioneer (SAST). We gradually lowered and raised the temperature at a rate of 0.5°C/hour and recorded the temperature at which test subjects entered and left the typical brumation posture (i.e., motion-less four-point stance with the nictitating membranes drawn across the eyes). These threshold temperatures were tightly associated with the corresponding field measurements both for the start (r = 0.97, t28 = 22.26, P < 0.0001, λ = 0.04 [0.00, 0.47]) and end of the inactive state (r = 0.98, t28 = 28.05, P < 0.0001, λ = 0.04 [0.00, 0.43]). Hence, we assessed the corresponding temperatures for all remaining species in the laboratory and estimated the brumation period based on the daily mean temperatures at the corresponding collection sites as retrieved from Chinese Meteorological Stations (http://www.lishi.tianqi.com) between 2012 and 2016. We defined the brumation period as the number of consecutive days in each year that remained below this threshold. For simplicity we determined the active rather than brumation period, starting with the first day that the mean daily temperature rose above the activity threshold and remained there for at least five consecutive days, and ending with the last day before the temperature dropped below the activity threshold and remained there until the end of the calendar year. The brumation period then represented the difference between the activity period and the total number of days in each calendar year. Across these five years, the measured temperature thresholds yielded highly repeatable species-specific estimates of the number of days below the activity range (R = 0.95 [95%CI: 0.93−0.96]), as determined by the rpt function in the rptR package (Stoffel et al. 2017) across all 116 species (Fig. S1; Table S1). Further, across the 30 species that were examined both in the lab and the field (see above), these predicted brumation periods were also correlated with the observed brumation periods in the field (r = 0.96, t28 = 18.03, P < 0.0001, λ = 0.05 [0.00, 0.48]; Fig. S8A), which themselves were highly repeatable between years within species (R = 0.98 [0.96−0.99]; Table S1).

Based on this data validation, we used for each species the mean brumation period predicted from our experimentally simulated temperature thresholds. However, to test for potential buffering effects of burrowing in the soil relative to the air temperatures reported by the meteorological stations, we also repeated these estimates by using more conservative thermal thresholds. Here, we restricted the putative brumation days to those with a reported air temperature of either 2°C or 4°C below the experimentally derived inactivity thresholds, simulating prolonged activity by seeking shelter in burrows. The 2°C threshold was based on a pilot study comparing direct measurements of air and burrow temperatures for four different burrows in each of five of our study species (burrow depths: 32.0 ± 3.2 to 121.0 ± 17.8 cm; Fig. S9). Across these species, the burrow-to-air temperature difference reached 1.03 ± 0.35°C to 2.45 ± 0.60°C in measurements around the peak of the brumation period (i.e., early January; Fig. S9). However, since these temporal snapshots were based on sites at relatively low elevation (≤320 m a.s.l.) due to accessibility of burrows during winter, we also used a second, more conservative buffer (4°C below activity range) for comparison. These temperature buffers shortened the predicted brumation periods to a varying degree between species (Fig. S8B); yet the predicted periods covaried strongly between the different temperature thresholds (all r > 0.90, t114 > 21.96, P < 0.0001, all λ < 0.01).

Phylogeny reconstruction

To reconstruct the phylogeny, we obtained the sequences of three nuclear and six mitochondrial genes from GenBank (for accession numbers and sequence coverage see Data S4). The three nuclear genes included the recombination-activating gene 1 (RAG1), rhodopsin (RHOD) and tyrosinase (TYR). The six mitochondrial genes were cytochrome b (CYTB), cytochrome oxidase subunit I (COI), NADH dehydrogenase subunits 2 and 4 (ND2 and ND4), and the large and small subunits of the mitochondrial ribosome genes (12S/16S; omitting the adjacent tRNAs as they were difficult to align and represented only a small amount of data). We aligned the sequences by multi-sequence alignment (MUSCLE) in MEGA v.10.2.2 (Tamura et al., 2013) before comparing possible nucleotide substitution models. The best substitution model, as determined by the function modelTest() in the R (R Core Team, 2022) package phangorn (Schliep, 2011) based on the corrected Akaike Information Criterion, AICc, was GTR+Γ+I for all genes except RHOD, for which HKY+Γ had stronger support. Using BEAUTi and BEAST v.1.10.4 (Suchard et al., 2018), we then constructed the phylogeny with unlinked substitution models, a relaxed uncorrelated log-normal clock, a Yule speciation process, and the best-supported nucleotide substitution models. We omitted time calibration due to a lack of fossil dates. We ran the Markov Chain Monte Carlo (MCMC) simulation for 55 million generations while sampling every 5,000th tree with a 10% burn-in. Most effective sample size (ESS) values by far exceeded 375 (i.e. all well above the recommended threshold of 200) for all but two tree statistics in the program Tracer v.1.7.2 (Rambaut et al., 2018), thus indicating satisfying convergence of the Bayesian chain and adequate model mixing. Finally, we generated a maximum clade credibility tree with mean node heights and a 10% burn-in using TreeAnnotator v.1.10.4 (Suchard et al., 2018), presented in Fig. S10.

Breeding conditions

To test if a prolonged brumation period reduces the time available for reproduction, thereby changing the level of competition over mates and fertilizations (Lüpold et al., 2017), we extracted the start and end dates of the breeding season from our field notes of concurrent studies on species-specific life histories. These data were available for 43 of our species (Data S3). We used recorded dates when the first and last clutches were observed in focal ponds as a proxy of mating activity, given that males release their sperm during oviposition in these external fertilizers. For each species, dates from at least two years were combined and averaged to obtain the mean duration of the breeding season.

We further recorded whether dense mating aggregations are typically observed in these species. We have previously shown that larger mating clusters, with multiple males clasping the same females, have a significant effect on the evolution of testis size due to the resulting competition among sperm for fertilization (Lüpold et al., 2017). Here, we had no detailed data on the sizes of aggregations and so were only able to code the typical presence or absence of aggregations as a binary variable (Data S3).

Finally, we used our direct estimates of species-specific population densities from our previous study (Lüpold et al., 2017) to test whether a shorter breeding season results in denser breeding populations. Although population density is a more direct measure than the occurrence of aggregations, such data were available for only eight of our species, each based on multiple populations per species (Lüpold et al., 2017). All these data were not necessarily derived from the same years or populations of our main dataset, but given the within-species repeatability in breeding populations (Lüpold et al., 2017) and in the duration of the breeding season (R = 0.88 [0.79−0.93]; Table S1), these differences should be relatively small compared to the interspecific variation and mostly introduce random noise.

Data analyses

We conducted all statistical analyses in R v.4.2.0 (R Core Team, 2022), using log-transformed data for all phenotypic traits, and for the CV in temperature among the ecological variables. To account for non-independence of data due to common ancestry (Pagel, 1999; Freckleton et al., 2002), we conducted phylogenetic generalized least-squares (PGLS) or phylogenetic logistic regressions (e.g., for occurrence of breeding aggregations), using the R package phylolm (Ho and Ané, 2014) and our reconstructed phylogeny. To account for variation around the species means, we bootstrapped for each model (at 100 fitted replicates) the standardized regression coefficients along with the phylogenetic scaling parameter λ and calculated their corresponding 95% confidence intervals. The λ values indicate phylogenetic independence near zero and strong phylogenetic dependence near one (Freckleton et al., 2002).

Unless stated otherwise, all PGLS models focusing on the relative mass of tissues as the response included snout-vent length (SVL) as a covariate in addition to the focal predictor variable(s). We chose SVL instead of body mass because it is the commonly used measure of body size in anurans and independent of seasonal fluctuations in tissues such as body fat, testes, or limb muscles. One exception, however, was the analysis of phylogenetically informed allometric relationships, for which we cubed SVL such that a slope of 1 equaled unity (isometry). For these allometric relationships we calculated ordinary (generalized) least-squares rather than reduced major-axis regressions, because their greater sensitivity to changes in the steepness, but lower sensitivity to changes in scatter, capture allometric slopes more adequately (Kilmer and Rodríguez, 2017).

To examine the covariation between different tissues across species, we first calculated pairwise partial correlations controlling for SVL and phylogeny. To this end, we calculated the phylogenetic trait variance-covariance matrix between the pairs of focal variable and SVL using the function phyl.vcv() in phytools (Revell, 2012) with λ = 1 (i.e. Brownian motion), which we then scaled into a correlation matrix using cov2cor() in the stats package (R Core Team, 2022). Using the resulting correlation coefficients rxy, rxz, and ryz, respectively, we then calculated the partial correlation coefficient rxy.z between the x and y variables of interest while accounting for SVL (z) following Crawley’s (Crawley 2007) equation: , with the associated t-statistics and 95% confidence intervals converted using standard conversion and the package effectsize (Ben-Shachar et al., 2020), respectively.

Since pairwise correlations do not necessarily capture more complex, multivariate allocation patterns, we used two additional approaches to explore how tissue sizes varied relative to others. Here, we focused on those four tissues that covaried with brumation duration or deviated from proportionate scaling with body size: brain, body fat, testes, and hindlimb muscles. Using the function acomp() in the R package compositions (van den Boogaart and Tolosana-Delgado 2008), we partitioned total body mass of each species into a five-variable Aitchison composition in a logistic geometry (van den Boogaart and Tolosana-Delgado, 2013), consisting of the proportional representation of the four focal tissues and the combined remaining body mass. Since the focal tissues constituted a size-independent fraction of the total body, the closed composition of this combined mass should be unbiased relative to body size but can instead reveal differential contributions of these tissues to their total in a multivariate context (Aitchison, 1982; Muldowney et al., 2001; van den Boogaart and Tolosana-Delgado, 2013). For phylogenetic correlations between these variables following the description above, we used centered log ratios obtained by the function clr() in the same package, which maintains the original variable structure. However, owing to the reliance on a full rank of the covariance in multivariate analyses, we used the ilr() function to project the D-part composition isometrically to a D−1 dimensional simplex (Aitchison, 1982), essentially representing the log ratios between the D parts. This multivariate object we subjected to a phylogenetic multivariate regression against brumation duration using the functions mvgls() and manova.gls() in the package mvMORPH (Clavel et al., 2015). For interpretation in the context of the original variable space, we back-transformed the coefficients using the ilrInv() function in compositions. In addition to this compositional data analysis, we also performed a phylogenetically informed principal component analysis on the same focal tissues as log-transformed species means, using the phyl.pca() function of the package phytools (Revell, 2012). Here, we primarily focused on the directions of the loading vectors relative to the principal components and one another to glean information on the correlations between the original variables in the principal component space. Finally, we performed phylogenetic confirmatory path analyses (von Hardenberg and Gonzalez-Voyer 2013; Gonzalez-Voyer and von Hardenberg, 2014) based on pre-specified candidate structural equation models, either to explore the direct and indirect effects of climatic variables on the duration of the breeding season or formation of breeding aggregations, or to better disentangle the different interrelationships between traits that could ultimately mediate the effect of brumation duration on brain and reproductive evolution. Using the R package phylopath (van der Bijl 2018), we examined the conditional independencies of each model, ranked all candidate models based on their C-statistic Information Criterion (CICc), and then averaged those with ΔCICc ≤ 2 from the top model (von Hardenberg and Gonzalez-Voyer, 2013). To avoid overparameterization of the path analysis on breeding parameters, which was based on only 43 species, we did not include testis mass (and so necessarily also SVL to control for body size) as additional variables in the same path models. However, we separately tested for correlated evolution using directional tests of trait evolution (Pagel, 1994; Revell, 2012). Based on (the weight of) the Akaike Information Criterion (AIC), we tested if changes in relative testis size and breeding parameters, respectively, were unilaterally dependent, mutually dependent, or independent (Pagel, 1994), using the fitPagel() function in the phytools package (Revell, 2012) with “fitDiscrete” as the optimization method and allowing all rates to differ (i.e., “ARD” model). Since these analyses are based on evolutionary transitions between binary states, we considered positive residuals of a log-log regression between testis mass and SVL as ‘relatively large testes’ and negative residuals as ‘relatively small testes.’ For the duration of the breeding season, we similarly split the distribution based on the mean duration, whereas aggregation formation was already coded as present or absent.

eLife Assessment

In this important paper, the authors report a link between brumation and tissue size in frogs, summarizing convincing evidence that extended brumation is associated with smaller brain size and increased investment in reproduction-related tissues. The research will be of broad interest to ecologists, evolutionary biologists, and those interested in global change biology. While the dataset involves significant field work and advanced statistical analyses, the manuscript would benefit from more explanation of the models, including why frogs are a good model in which to address these questions, and from general improvement in the structure and conciseness.

Reviewer #1 (Public Review):

The authors have conducted lots of field work, lab work and statistical analysis to explore the effect of brumation on individual tissue investments, the evolutionary links between the relative costly tissue sizes, and the complex non-dependent processes of brain and reproductive evolution in anuran. The topic fits well within the scope of the journal and the manuscript is generally written well. The different parameters used in the present study will attract a board readership across ecology, zoology, evolution biology, and global change biology.

Reviewer #2 (Public Review):

The authors set out to show how hibernation is linked to brain size in frogs. If there were broader aims it is hard to decipher them. The authors present an extremely impressive dataset and a thorough set of cutting-edge analyses. However not all details are well explained. The main result about hibernation and brain size is fairly convincing, but it is hard to think of broader implications for this study. Overall, the manuscript is very confusing and hard to follow.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

In this important paper, the authors report a link between brumation and tissue size in frogs, summarizing convincing evidence that extended brumation is associated with smaller brain size and increased investment in reproduction-related tissues. The research will be of broad interest to ecologists, evolutionary biologists, and those interested in global change biology. While the dataset involves significant field work and advanced statistical analyses, the manuscript would benefit from more explanation of the models, including why frogs are a good model in which to address these questions, and from general improvement in the structure and conciseness.

We highly appreciate your positive assessment and that you considered our paper important and convincing.

Reviewer #1 (Public Review):

The authors have conducted lots of field work, lab work and statistical analysis to explore the effect of brumation on individual tissue investments, the evolutionary links between the relative costly tissue sizes, and the complex non-dependent processes of brain and reproductive evolution in anuran. The topic fits well within the scope of the journal and the manuscript is generally written well. The different parameters used in the present study will attract a board readership across ecology, zoology, evolution biology, and global change biology.

Thank you for your positive and supporting feedback.

Reviewer #2 (Public Review):

The authors set out to show how hibernation is linked to brain size in frogs. If there were broader aims it is hard to decipher them. The authors present an extremely impressive dataset and a thorough set of cutting-edge analyses. However not all details are well explained. The main result about hibernation and brain size is fairly convincing, but it is hard to think of broader implications for this study. Overall, the manuscript is very confusing and hard to follow.

Thank you for your compliments on our paper. As for your concerns, we have greatly revised our paper and, as we hope, improved its clarity. We have also added a few sentences to the conclusions to draw attention to potentially broader implications. Specifically, we describe how the focal traits of our study may all be affected by climate change. Differential constraints in necessary investments could be one of several reasons for the varying resilience to climate change between species in the same habitat.

Reviewer #1 (Recommendations For The Authors):

There are no issues on the availability of data and code.

Thank you.

Line 15: in the author contribution section, it seems that C.L.M. and J.P.Y are not in the author list.

These two authors are not part of this study. This was a mistake.

Line 24: I don't think it is vital or logical to address or compare too much on birds or mammals, which are not the focused taxa of the present study. Instead, it is better to clarify the reason why frogs and toads are ideal model taxon to this study.

The reason for comparisons with birds and mammals was that all hypotheses related to the various trade-offs tested here had been developed in these taxa. One of the points of our paper was that these needed validation beyond the two taxa, in addition to being tested against one another (each prediction had been developed in a specific group and typically in isolation of all other hypotheses).

Line 25-26: as the authors are shooting for eLife, as a general journal, it is not essential to provide the detailed methods in the abstract. But I think the authors need to strengthen the novelty of the work in the field here.

The strength of our study was that all traits were measured directly in our species, including estimates of hibernation duration. Prior studies used various proxies, categorial classification or datasets assembled from multiple sources. To us, this seemed like a sufficiently important advance in the field to mention it, but considering the reviewer’s comment, we have now removed it.

Line 28: "protracted brumation reduces brain size and instead promotes reproductive investments", as a correlative study, it is much more precise to change this sentence to a similar description as "protracted brumation is negatively correlated with brain size but is positively correlated with reproductive investments" here and related statements throughout the whole text.

We agree that, strictly speaking, a path analysis can only point toward possible causality and not provide hard evidence as experimental manipulation might. The wording may have been a bit too strong here in our attempt to minimize wordiness and because all our analyses combined very strongly pointed in this direction. However, we have now changed this as suggested even though it now reads almost as if we had done no more than conducting a simple correlation. We have further paid attention to the wording of our interpretations throughout the paper.

Line 32-33: it needs a bigger ending linking your main findings with the implication in understanding species response to the sustained environment change.

We have reworded the ending of the abstract to: “Our results provide novel insights into resource allocation strategies and possible constraints in trait diversification, which may have important implications for the adaptability of species under sustained environmental change.”

Line 63-68: this sentence is too long to understand and please simplify it.

We have split the sentence into two sentences.

Line 125-130: it is known that there are various frog reproductive modes (Crump et al. 2015) such as trade-offs between clutch size and egg size, different number of breeding during one year, etc. These different reproductive forms may also influence the brain size evolution with food availability and seasonal variations. Please clarify it.

Yes, anurans do have varying reproductive modes, but to us, there is no a priori reason to assume that such variation would have a direct effect on brain evolution. Rather, in our opinion, different reproductive modes would have indirect effects by affecting the environment in which reproduction occurs. For example, larvae developing under different environmental conditions (substrate, larval density, egg provisioning etc.) might affect developmental trajectories that could influence how resources are available and allocated to different organs, including the brain. Alternatively, reproductive modes could influence the choice of environment for reproduction, thereby possibly affecting mating strategies and ultimately trait investments associated with these strategies. Given we were asked to shorten our paper, we believe that ‘environmental effects’ remains broad enough to encompass such variation, thereby not necessitating disentangling the different, and likely primarily indirect, ways that reproductive modes could be linked to brain evolution. However, if the reviewer would find it important to go into such detail in the paper, we will be happy to do so.

Line 186-187: it is necessary to mention here that the authors also conducted sensitivity analyses to apply 2{degree sign}C or 4{degree sign}C below their experimentally derived as thresholds to test the robustness of the results to data uncertainty.

We have added “(details on methodology and various sensitivity analyses for validation in Material and Methods)” to indicate the different types of sensitivity analyses, which included more than simply 2 or 4°C difference.

Line 188: please change "In phylogenetic regressions" to "after controlling for phylogenetic autocorrelation/pseudo-replication" or similar sentence here.

Our focus here was the phylogenetically informed GLS model rather than phylogenetic control itself. In the latter case, it would still not be clear what type of model was conducted with such phylogenetic control. To avoid any shorthand, we have reworded for more precision: “We employed phylogenetic generalized least-squares (PGLS) models, …”

Line 177-287: please provide the exact variance explained by different predictor variables in brumation duration, individual tissue investments, and brain evolution. I also suggest that the authors need consider conducting multi-model inference-based model averaging analysis to test the relative importance of different variables. In addition, the present analyses did not include the interaction terms among variables, which may be more important than the effect of each individual factor.

There may be some misunderstanding as these models represent separate analyses for each predictor as indicated by the associated λ values (never more than one value per model). We conducted separate models to determine which variables might even play a role in explaining variation in the corresponding response variables. Based on relevant predictors, we then conducted path analyses rather than general multi-predictor analyses. The relative effect sizes are represented by the correlation coefficients (r values) in the tables.

Reviewer #2 (Recommendations For The Authors):

Why exactly are the pairwise comparisons positively correlated (fig. S5) and then negatively correlated (fig. 3). What is actually driving this difference? For the phylogenetic path analyses 26 candidate models are chosen without explanation. What theory or hypotheses are these based on?

We assume the reviewer is referring to the brain-body fat association. The two ‘pairwise’ analyses they mention were not the same. The correlation in Fig. S5 was a standard (albeit phylogenetically informed) partial correlation between the two focal tissues, controlling for SVL. By contrast, as described when introducing the analyses, negative associations were derived when additionally controlling for testes and hindlimb muscles, all of which deviated from isometry against body size. Here, the total mass of the four main tissues was divided by their proportional contribution to that mass in each species, then standardized for comparison across species. Since the total mass of these four tissues scaled directly with body size, larger-bodied species did not invest a proportion of their body to these tissues than smaller-bodied species, thus essentially rendering body size irrelevant for this analysis. However, the relative representation of the four traits changed between species such that more resources devoted to body fat was associated with a smaller brain, hence a negative relationship. Similarly, the multivariate analysis as well as the PCA also suggested similar trends when all four tissues were considered rather than purely pairwise comparisons.

Regarding the second comment: We indeed used 28 pre-defined predictions for our larger path analysis.

The authors haven't really provided much additional context either, and the discussion is almost entirely a rehash of the results section. I can't see the analysis code but this may be of use to people performing similar analyses.

It is true that the traits and core message of the Discussion relate directly to our results, but we believe that our Discussion provides the essential biological context to our findings and to how they are connected. We tried not to go on tangents or too much speculation as the many results provided enough material to discuss, with several different ways that we expanded the prior state-of-the-art in the field. However, we have now expanded the concluding paragraph to place our findings in the context of climate change, given that this could affect anurans and the different traits examined in many ways that are directly related to the current study. Yet, we decided to keep this short because such extrapolation of our findings

We indeed held off making the code available to the public in case dramatic changes to the paper were requested by the reviewers. However, it will be published.

Additional recommendations from the Reviewing Editor:

One of the reviewers and I found the text a little difficult to follow. I suggest simplifying the paper by being more concise. For example, the introduction could be shortened into a 3-4 paragraphs of relevant text without overwhelming the reader. One of the reviewers wanted a better explanation of statistical models and I agree. The discussion could benefit from some structure - consider adding subheadings that would guide the reader as to the topic. Finally, the figures are difficult to see and should be made larger. For example, the graphs in Figure 1c could be on a panel below A and B so that readers can interpret the graph. In Figure 3 - the legend is far too small - please put above or below the graphs. In summary - I hope you consider a major re-write that would strengthen the accessibility of your paper to a broad audience.

We have substantially shortened the paper despite adding further details on models and a broader context to the Discussion. We also condensed the Introduction to about two thirds of the original word count. However, we did not think that shortening it even further or splitting it into 3-4 paragraphs would improve readability. We still considered it important to introduce with sufficient context all major hypotheses that were tested against one another, provide at least some information on what was or was not known about the evolution of the focal traits and their links to one another or the environmental variables. We also found it important to touch on the differences between our study organisms and those typically studied in the context of hibernation or brain evolution, as this could affect the predictions. Given the number of hypotheses and traits, cutting the number of paragraphs would have meant merging some of them into very long ones, which we did not consider helpful.

We further added short subheadings to the Discussion and adjusted the figures as requested.

Version 2

Abstract

Abstract

Environmental seasonality can promote the evolution of larger brains through cognitive and behavioral flexibility but can also hamper it when temporary food shortage is buffered by stored energy. Multiple hypotheses linking brain evolution with resource acquisition and allocation have been proposed for warm-blooded organisms, but it remains unclear how these extend to cold-blooded taxa whose metabolism is tightly linked to ambient temperature. Here, we integrated these hypotheses across frogs and toads in the context of varying brumation (‘hibernation’) durations and their environmental correlates. We showed that protracted brumation covaried negatively with brain size but positively with reproductive investment, likely in response to brumation-dependent changes in the socio-ecological context and associated selection on different tissues. Our results provide novel insights into resource allocation strategies and possible constraints in trait diversification, which may have important implications for the adaptability of species under sustained environmental change.

Introduction

Seasonal food scarcity challenges animal energy budgets. A positive or less negative energy balance across seasons can be achieved by a more constant net energy intake than predicted solely by food abundance (Sol, 2009), or by investing less in costly organs (Heldstab et al., 2018). A link between both strategies is the brain: A relatively large brain can improve cognitive ability and behavioral flexibility (Benson-Amram et al., 2016; Lefebvre et al., 2004; Reader and Laland, 2002; Sol et al., 2005), enabling animals to effectively locate diverse and dispersed food sources to buffer environmental fluctuations in seasonal habitats (‘cognitive buffer hypothesis’) (Allman et al., 1993; Sol, 2009; van Woerden et al., 2012). However, brain tissue has high metabolic costs (Aiello and Wheeler, 1995; Lukas and Campbell, 2000; Mink et al., 1981) that may not be temporarily reducible (Mink et al., 1981), constraining brain size evolution under periodic food scarcity (‘expensive brain hypothesis’) (Isler and van Schaik, 2009).

Food scarcity can also affect physiological responses. For example, a longer digestive tract may permit more efficient resource uptake during a short active period and be favored by selection in species with prolonged hibernation (Sibly, 1981). The evolution of the digestive tract could thus parallel that of the brain. By contrast, there could be an evolutionary trade-off between the two organs (‘expensive tissue hypothesis’; Aiello and Wheeler 1995; also see Isler and van Schaik (2006) for a more general ‘energy trade-off hypothesis’). Furthermore, physiological buffering often involves a seasonal reduction in metabolic rate or activity (e.g., hibernation), with energy drawn from stored fat reserves (Heldstab et al., 2016). Buffering lean periods by fat stores and reduced activity contrasts with cognitive abilities to cope with food scarcity, suggesting differential investment in adipose and brain tissue between species in highly seasonal environments (‘fat−brain trade-off hypothesis’; Heldstab et al., 2016; Navarrete et al., 2011). Finally, allocating fat stores to sustained brain function throughout prolonged hibernation might compete with investments in other tissues (Isler, 2011; Isler and van Schaik, 2009).

Species that breed soon after hibernation, such as some mammals (Place et al., 2002; Psenner, 1957) and many amphibians (Fei and Ye, 2001; Wells, 1977), face further unique challenges, as their reproductive tissue must regrow before emergence when stored resources are most limited (Isler, 2011; Isler and van Schaik, 2009). Testes, however, may be subject to intense selection by sperm competition resulting from female multiple mating (Lüpold et al., 2020), a widespread phenomenon where males cannot monopolize their mates (Lüpold et al., 2014). Indeed, across anurans (frogs and toads) that often breed in small water bodies, males invest relatively more in their testes and less in their forelimbs (used in pre-mating competition) as population density increases (Buzatto et al., 2015; Lüpold et al., 2017). If a shorter active period leads to more synchronized breeding and thus increases the risk of sperm competition or sperm depletion (Vahed and Parker, 2012), selection for relatively larger testes would be stronger where fat stores need to last longer, affecting resource demands and allocation during hibernation. Including reproductive investments and breeding patterns in studies of allocation trade-offs in response to hibernation and environmental seasonality would thus seem critical but remains to be done, particularly in the context of brain evolution.

The opposing selection pressures on brain or gonad size (i.e., cognitive or fitness benefits versus metabolic costs), varying degrees of seasonality and diverse strategies for buffering periodic food scarcity (e.g., cognitive versus physiological) between species render environmental fluctuations an ideal context for studying brain size evolution. The different hypotheses for the coevolution of brain size with other organs were developed separately for different mammalian or avian taxa, the two vertebrate classes with the largest brains relative to body size (Jerison, 1973). These hypotheses have yet to be directly tested against one another in a single taxon and ideally in the immediate context of seasonal activity, considering the extent, rather than the mere presence/absence, of hibernation. This last point is important because ‘hibernation’ can range between brief inactive bouts and long dormancy, with different energetic and life-history constraints. Furthermore, understanding the generality of the patterns reported in mammals or birds needs to be validated in other taxa, ideally with smaller brains and different energy demands. Such a generalization would help contextualize brain evolution in the two largest-brained taxa in relation to selection on encephalization and metabolic constraints.

A particularly suitable system is presented in ectothermic (‘cold-blooded’) species whose metabolism and activity are tightly linked to ambient temperature (Wells, 2007), possibly leading to stricter physical boundaries to behavioral flexibility than in the endothermic (‘warm-blooded’) mammals or birds. Compared to the divergent groups of mammals, for which most hypotheses on brain evolution have been proposed, anurans are also more homogeneous in body size and shape, diet and locomotion (Kardong, 2019), while still showing strong environmental effects on brain, reproductive and other investments (Liao et al., 2022; Lüpold et al., 2017; Wells, 2007). Being less influenced by Bauplan and lifestyle, evolutionary trade-offs with brain size could thus be easier to isolate in anurans than in mammals.

Here, we studied across 116 anuran species how varying ‘hibernation’ periods and their environmental correlates affect male investments in brain, testes, body fat, limb muscles, and the main visceral organs (see Fig. 1).Unlike mammals, anurans do not actively depress their metabolism but rather stop activity when ambient temperatures drop below the activity range (referred to here as ‘brumation’ for distinction; Pinder et al. 1992; Wells 2007). However, in both taxa, hibernators use fixed energy stores across sequential investments, albeit at a lower metabolic rate (Staples, 2016), whilst non-hibernators can partly replenish resources but need more food even when it is scarce (Heldstab et al., 2016). These contrasting strategies may be associated with a varying cost-benefit balance between organs, including the brain.

Allometric and seasonal variation in the species-specific tissue masses. A: Allometric slopes between the mass of each tissue and cubed snout-vent length (SVL3) so that proportionate scaling follows a slope of 1 on a log-log scale. Each point represents a species-specific mean value in breeding condition (N = 116). Relationships deviating from proportionate scaling (based on bootstrapped 95% confidence intervals) are highlighted in blue (steeper than unity) or red (shallower than unity). B: Mean percent change with 95% confidence interval for body mass and each individual tissue of 50 anuran species with data from both shortly before and after brumation (= breeding), based on absolute tissue masses between stages and log-transformed to maintain symmetry and additivity (Törnqvist et al., 1985): log(post-brumation / pre-brumation) × 100. The transparent grey dots depict species-specific values. C: Relationship between brumation period and percent mass change in the amount of body fat. D: Relationship between brumation period and percent mass change in testis mass. Each point indicates a species.

Although anurans halt general activity during brumation, this does not mean complete brain inactivity. Several species move in their burrows or underwater hibernacula in response to changes in soil temperature or oxygen concentration (Holenweg and Reyer, 2000; Stinner et al., 1994; van Gelder et al., 1986), or when disturbed (Niu et al., 2022; Tattersall and Ultsch, 2008). Moreover, brumating frogs can renew more brain cells than active frogs, possibly to prevent brain damage (Cerri et al., 2009), and this could increase with brain size. Larger brains could also be less tolerant to low oxygen conditions (Pinder et al., 1992; Tattersall and Ultsch, 2008; Wells, 2007) if the findings from other ectotherms extend to anurans (Sukhum et al., 2016). Overall, maintaining a relatively larger brain while brumating may entail higher costs that could constrain brain evolution compared with species with short or no seasonal inactivity.

Instead of the typical indirect proxies or scores (Heldstab et al., 2018, 2016; Luo et al., 2017; Navarrete et al., 2011), we quantified seasonal changes in tissue size and estimated brumation duration as the time that temperatures were continuously below the species-specific activity threshold. These estimates aligned with periods of inactivity recorded during field surveys, and were robust across multiple sensitivity analyses (e.g., considering temperature buffering by underground burrows). We then tested the different hypotheses on brain evolution and further integrated variation in breeding context and reproductive evolution in this resource allocation framework. By examining ectothermic species with diverse energy demands and brain sizes, our findings provide a crucial context for understanding brain evolution and resource allocation.

Results

Determinants of brumation duration

For a mean (±SD) of 3.41 ± 0.95 males in each of 116 anuran species, we combined experimentally determined thermal activity thresholds with multi-year temperature fluctuations at their collection sites to estimate the species-specific brumation periods (details on methodology and various sensitivity analyses for validation in Material and Methods). These periods averaged between 0.6 ± 0.5 and 250.5 ± 16.7 days across the five years examined, with high repeatability within species (R = 0.95 [95%CI: 0.93, 0.96]; Fig. S1; Table S1). These brumation periods largely fell into three groups with ≤9 (N = 22 species), 18−27 (N = 3), and ≥47 days of expected brumation (N = 91), respectively. As the ground microclimate may buffer some of the fluctuations in air temperature and frogs can endure short cold spells without dormancy, we conservatively considered those 25 species with ≤27 days below their experimental temperature threshold unlikely to show any sustained brumation.

We employed phylogenetic generalized least-squares (PGLS) models (Freckleton et al., 2002; Ho and Ané, 2014) to dissect the factors driving brumation duration. Our analyses indicated that the brumation period increased with the latitude and elevation of study sites, along with the extent of annual temperature fluctuations (r ≥ 0.42, t114 ≥ 5.00, P < 0.001; Table S2). In contrast, brumation duration was inversely correlated with the annual mean temperature and precipitation, and dry season length (r < −0.29, t114 < −3.24, P < 0.002), but not significantly covarying with longitude or annual fluctuations in precipitation (r < 0.09, t114 < 0.95, P > 0.34; Table S2). Except for the duration of the dry season, these results persisted for those 91 species with a high probability of sustained brumation (see above; Table S3). Among these species, those from cooler and more seasonal climates exhibited lower entry and emergence temperatures from their inactive state (Table S4), suggesting heightened cold tolerance to maximize their active period.

Impact of brumation on tissue investments

To investigate the repercussions of brumation on individual tissue investments, we used the same males from our 116 species. In separate sets of PGLS models, neither snout-vent length (SVL) nor body mass covaried with brumation duration or any other environmental variable (|r| ≤ 0.15, |t114| ≤ 4.03, P ≥ 0.11; Table S5), except for a weak, non-statistically significant trend toward reduced body mass at higher elevations (r =−0.18, t114 = −1.91, P = 0.06, phylogenetic scaling parameter λ = 0.94 [95%CI: 0.85, 1.00]). The different tissues increased with body size, albeit with different allometric slopes. Body fat and limb muscles showed steeper scaling (all β ≥ 1.10 [1.03, 1.18]), brain size exhibited a shallower slope (β = 0.49 [0.44, 0.54]; Table S6; Fig. 1A), and the remaining tissues did not deviate from proportionate scaling (i.e., 95%CI including 1.00; Table S6; Fig. 1A). Hence, the evolution of brain size appears to be more constrained than that of other organs when selection favors larger body size.

Given these distinct allocation patterns between brain and other tissues, we next tested whether extended brumation constrains brain size evolution, akin to suggestions for mammals based on presence/absence of hibernation (Heldstab et al., 2018). In PGLS models, absolute brain size was independent of brumation duration (r = −0.10, t115 = −1.08, P = 0.28, λ = 0.89 [0.73, 0.97]), but accounting for SVL as a body size proxy, species with prolonged brumation had relatively smaller brains, quantified during breeding (partial r, rp = −0.31, t113 = −3.50, P < 0.001, λ = 0.35 [0.00, 0.61]; Fig. 2A, Table S7) or pre-hibernation (N = 50 species means based on 2.64 ± 0.94 males each: rp = −0.51, t47 = −4.03, P < 0.001, λ = 0.00 [0.00, 0.56]; Table S7). This trend was distinct from body size changes in response to brumation, as SVL remained independent of brumation duration (Table S5). On average, brumating species tended to possess relatively smaller brains compared to those less likely to experience prolonged brumation (Table S8), aligning with mammalian studies utilizing hibernation presence/absence (Heldstab et al., 2018). This trend extended to those 91 species categorized as brumating for some period (Table S9), reinforcing the link between brumation and brain evolution beyond coarse binary classification. This pattern further persisted when recalculating brumation periods using conservative thresholds of 2°C or 4°C below their experimentally derived thresholds, simulating shelter buffering (details and validation in Material and Methods; Table S10).

Effects of brumation duration on the relative tissue sizes. Relationships between brumation duration and the relative mass of the brain (A), body fat (B), and testes (C) across males of 116 anuran species in breeding (post-brumation) condition. All axes are controlled for the snout-vent length and phylogeny.

Species displaying extended brumation periods further exhibited relatively more body fat (rp ≥ 0.25, t113 ≥ 2.72, P ≤ 0.008; Fig. 2B) and, particularly during breeding, had relatively larger testes (rp = 0.36, t113 = 4.06, P < 0.001, λ = 0.77 [0.40, 0.90]; Fig. 2C), along with relatively smaller hindleg muscles (rp = −0.22, t113 = −2.37, P = 0.02, λ = 0.22 [0.00, 0.51]; Tables S7). These patterns generally held true when using the presence/absence of brumation (Table S8), buffered temperature fluctuations (Tables S10), or excluding the 25 species unlikely to brumate (except for the non-significant effect on body fat; Table S9). Other tissue sizes remained independent of brumation (Tables S7 to S10) and showed not significant changes between sampling periods (Fig. 2).

Comparing pre-and post-brumation males, averaged across all species, we found a roughly 50% reduction in fat tissue and a 100% increase in testis size, indicating resource depletion and testicular regrowth during brumation, respectively (Fig. 1B). Only brain size deviated from zero among the remaining tissues, but this change was minimal compared to the body fat and testes, and within the range of many unchanged tissues. Thus, the biological significance of this putative increase in brain size during brumation remains uncertain, possibly reflecting general differences between sampled individuals. We will thus refrain from further interpretation. Among the two tissues with considerable change, the extent of fat depletion increased significantly with the brumation period (r = −0.36, t48 = −2.68, P = 0.01, λ = 0.00 [0.00, 0.55]), as did that of testis regrowth (r = 0.39, t48 = 2.89, P = 0.006, λ = 0.82 [0.16, 0.98]; Fig. 1C).

To examine whether the increase in relative testis size with prolonged brumation might be mediated by a shorter, more synchronized mating season (Wells, 2007), we tested for links between brumation duration and different breeding parameters. Prolonged brumation notably shortened the breeding season (r = −0.57, t41 = −4.47, P < 0.001, λ = 0.00 [0.00, 0.38]; Fig. S2A), exerting a stronger effect than climatic variables (Table S11), particularly when considered together (Table S12). Hence, the effect of these climatic variables may be mediated by brumation. Furthermore, a shorter breeding season increased the probability of dense breeding aggregations (phylogenetic logistic regression: N = 42, z = −3.03, P = 0.002, α = 0.02; Fig. S2B). That brumation might mediate climatic impacts on breeding aggregations through the duration of the breeding season was also supported by a phylogenetic confirmatory path analysis (Gonzalez-Voyer and von Hardenberg, 2014; von Hardenberg and Gonzalez-Voyer, 2013; Fig. S3; Table S13). Finally, when combining these data with published data on the density of breeding populations (Lüpold et al., 2017; N = 8 species overlapping), a trend emerged toward higher mean population densities in species with shorter breeding seasons (r = −0.69, t6 = −2.37, P = 0.06, λ = 0.00 [0.00, 1.00]), albeit with a small sample size (Fig. S2C).

To explore possible causal links between breeding parameters and relative testis size, we employed directional tests of trait evolution (Pagel, 1994; Revell, 2012), assessing whether changes in two binary traits are unilaterally or mutually dependent, or independent (Pagel, 1994). To this end, we converted our continuous to binary variables (see Material and Methods). Relating small/large testes to short/long breeding seasons, the best-supported scenario based on the Akaike Information Criterion (AIC) was independent evolution. Yet, the model with changes in relative testis size dependent on those in the breeding season found similar support (ΔAIC = 0.83, wAIC = 0.33 compared to independent model with wAIC = 0.50; Fig. S4A), differing from the remaining two models with clearly higher AIC scores (ΔAIC ≥ 3.40, wAIC ≤ 0.09). In another directional test, between relative testis size and aggregation formation, the independent model was the best-supported (wAIC = 0.66), followed by the scenario of increased relative testis size in response to aggregation formation (Fig. S4B), albeit above a ΔAIC cut-off of 2 (ΔAIC = 2.38, wAIC = 0.20), with the remaining models being clearly less supported (ΔAIC ≥ 3.64, wAIC ≤ 0.11). Hence, it is at least possible that breeding conditions mediate the positive relationship between brumation duration and relative testis size in our relatively small sample of species, whilst a response of breeding conditions to variation in relative testis size is clearly rejected by our analyses.

Covariation between tissues

As all tissues rely on the same finite resources, their responses to brumation are likely interconnected. Pairwise partial correlations, controlling for SVL and phylogeny, revealed positive covariation or non-significant associations among all tissue masses (Fig. S5). As such, our data do not support the expensive tissue (Aiello and Wheeler, 1995) or the more general energy trade-off hypotheses (Isler and van Schaik, 2006), which predict brain size trade-offs with the digestive tract or other costly organs, respectively. However, since brain size differed from fat, hindlimb muscles and testes in allometric relationships and responses to brumation, pairwise correlations may not capture more complex allocation. To examine the relative investments in these four most informative tissues simultaneously, we represented total body mass as proportions of these tissues and the remaining mass, generating a five-variable compositional dataset (van den Boogaart and Tolosana-Delgado, 2013). The combined mass of all four focal tissues scaled proportionately with body size (allometric β = 1.02 [0.96, 1.07], λ = 0.01 [0.00, 0.11]), confirming size-independent proportion of the total resources allocated to the four focal tissues combined. However, the resource distribution among these tissues varied considerably between species. Pairwise correlations between the four focal tissues, transformed to centered log ratios (van den Boogaart and Tolosana-Delgado, 2008) and controlling for phylogeny, showed brain mass to covary negatively with fat and testis mass, whilst testis mass covaried negatively with hindlimb muscle mass but not with body fat (Fig. 3A-D; Table S14).

Effects of brumation duration on the relative tissue sizes. Panels A-D depict the phylogenetic correlations (shown as phylomorphospace plots; Revell, 2012) between the relative masses of (A) brain and body fat, (B) brain and testes, (C) testes and body fat, and (D) testes and hindlimb muscles, respectively, across the 116 species (results in Table S14). The relative tissue masses represent the centered log ratios of the compositional data, and the lines connect the nodes of the underlying phylogeny, indicating that phenotypic correlations are not simply the result of phylogenetic clustering. The correlation coefficients and 95% confidence intervals are indicated. The loadings from a phylogenetic principal component analysis (Revell, 2012) on the same variables are also mapped as vectors onto biplots between (E) the first and second or (F) the second and third principal components. In all panels, the point colors reflect the species-specific brumation periods (see legend in panel A). Generally, where brumation was relatively shorter or absent, species also tended to have relatively larger brains, less body fat and smaller testes, respectively, consistent with the univariate analyses (Fig. 2).

To further examine the effect of brumation duration on all five variables simultaneously, we conducted a phylogenetic multivariate regression analysis (Clavel et al., 2015) on the same compositional data, but now transformed to isometric log ratios as recommended for multivariate models (van den Boogaart and Tolosana-Delgado, 2013, 2008). Brumation duration significantly affected anuran body composition (Pillai’s trace = 0.33, effect size ξ2 = 0.30, P = 0.001). The back-transformed coefficients of body fat (0.23) and testis mass (0.26) exceeded the expected coefficient of 0.20 if brumation had no effect, whilst brain mass, hindlimb muscles and the rest of the body had lower coefficients (0.16, 0.17, and 0.18, respectively; also see Fig. 3).

Finally, a phylogenetically informed principal component analysis (Revell, 2012) confirmed the negative associations of brain size with body fat and testis mass, and that of testis mass with hindlimb muscles (Fig. 3E,F, Table S15). Here, the first three principal components (PC1 to PC3) explained 84.7%, 8.0% and 5.1% of total variance, respectively. Although PC2 and PC3 explained a relatively small proportion of the total variance, they separated the different tissues. PC2 loaded mainly by brain size (0.30) and testis mass (−0.44), and PC3 by brain size (0.26) and body fat (−0.31). Biplots indicated negative associations between the vectors of the same traits as above within the multivariate trait space (Fig. 3F). Further, brumation duration covaried negatively with PC2 (r = −0.53, t114 = −6.59, P < 0.0001, λ = 0.74 [0.49, 0.92]), consistent with reduced brain size and increased in testis mass towards longer brumation, but it was not associated with PC3 (r = −0.12, t114 = −1.30, P = 0.20, λ = 0.41 [0.00, 0.70]).

Direct and indirect effects revealed by path analysis

To untangle the evolutionary links between tissue sizes, and to test competing brain evolution hypotheses in the context of brumation, we conducted a phylogenetic confirmatory path analysis (Gonzalez-Voyer and von Hardenberg, 2014; von Hardenberg and Gonzalez-Voyer, 2013) with 28 predetermined path models (Fig. S6; Table S16). The averaged model (Fig. 4) confirmed the negative effect of prolonged brumation on relative brain size (β = −0.15 [−0.22, −0.07]), along with direct (β = 0.16 [0.07, 0.26]) and indirect positive effects on relative testis size (Fig. S7). These effects on testis size were mediated by the relative amount of adipose tissue, which increased with brumation duration (β = 0.13 [0.06, 0.20]) and digestive tract size (β = 0.51 [0.37, 0.65]), and in turn positively influenced relative testis size (β = 0.39 [0.18, 0.59]).

Results of the averaged phylogenetic path model. Visual representation of the average phylogenetic path model across 116 anuran species. Arrows reflect the direction of the path, with their widths being proportional to the standardized regression coefficients and colors indicating the sign (blue = positive, red = negative). Paths with 95% confidence intervals excluding 0 (i.e., arrows highly probable) are drawn as solid arrows, all others as dashed, semi-transparent arrows. For simplicity and to avoid over-parameterization, other organs were omitted in path models as they showed little covariation with brumation duration or brain size. All phenotypic traits were log-transformed, and all variables were controlled for body size via additional paths from log SVL. Although SVL had a strong effect on all variables (all β > 0.37), its thick blue arrows to each box are omitted in this figure only for visual clarity, but all path coefficients are presented with their 95%CI in Fig. S7, with further details in Fig. S6 and Table S16.

Discussion

Our study on anurans, with validated brumation periods and direct measures of expensive tissues, provides novel insights into brain and reproductive evolution in ‘cold-blooded’ organisms exposed to environmental seasonality. We found that species with longer brumation exhibited relatively smaller brains and allocated greater fat reserves primarily in reproduction, possibly due to the shorter breeding season with its socio-ecological implications.

Environmental and resource considerations in brain evolution

We demonstrated that species in cooler and more seasonal climates were more cold tolerant, thereby likely optimizing their active period. Yet, low temperatures impair foraging and digestion in ectotherms (e.g., Fontaine et al., 2018; Riddle, 1909), such that high seasonality may lead to longer brumation and smaller brains. These results confirm that, unlike birds (Sol, 2009) and some mammals (van Woerden et al., 2010), challenging and unpredictable environmental conditions select for physiological rather than cognitive buffering in anurans (Luo et al., 2017). Supporting a large brain may not be sustainable without continued resource intake, or larger brains could be less tolerant to hypoxic conditions during brumation (Sukhum et al., 2016). However, selection for relatively larger brains may also simply be stronger in species with longer active (and short brumation) periods owing to extended cognitive benefits such as predator evasion (Kotrschal et al., 2015; Liao et al., 2022) or exploitation of better and more diverse food sources (Jiang et al., 2022; Lefebvre et al., 1997).

In pairwise comparisons, the relative sizes of the tissues examined here, including the brain, were generally positively correlated. These results reject both the expensive tissue and energy trade-off hypotheses (Aiello and Wheeler, 1995; Isler and van Schaik, 2006), which predict trade-offs of brain size with the size of the digestive tract or other costly organs, respectively. This lack of support in anurans aligns with a previous report in mammals (Navarrete et al., 2011) despite their smaller brains and vastly different ecology and physiology, including a lower metabolic rate and largely lacking physiological thermoregulation. When focusing jointly on the four tissues (brain, body fat, testes, hindlimb muscles) that covaried with brumation duration, however, relative brain size covaried negatively with the relative mass of both fat tissue and testes, supporting the fat−brain trade-off (Navarrete et al., 2011) or expensive sexual tissue hypotheses (Pitnick et al., 2006), respectively.

Brumation and fat tissue

Species with longer brumation periods exhibited relatively more total body fat and a higher degree of its depletion, supporting the hypothesis that anurans buffer lean periods by metabolizing stored fat (Huang et al., 2020; Luo et al., 2017). Although adipose tissue may not itself be metabolically expensive, transporting it adds costs to locomotion, particularly when jumping away from predators (Moreno-Rueda et al., 2020) or climbing trees compared to moving horizontally on land or in water (Alexander, 2003; Hanna et al., 2008). Consistent with this notion, arboreal species tended to be leaner compared to (semi)aquatic or terrestrial species (Table S17), controlling for brumation duration and relative brain size, both of which we had shown to covary with body fat (Figs. 2 and 3).

Brumation and testis evolution

Species with prolonged brumation also had relatively smaller hindleg muscles and larger testes. The negative relationship between hindleg muscle mass and brumation duration may be linked to more movement during a longer active period, including predator evasion (Liao et al., 2022; Marchisin and Anderson, 1978). Larger testes may result from a shorter breeding season, leading to denser and more synchronous mating activity (Wells, 2007), as suggested by our path analysis. Breeding aggregations increase male-male competition over fertilization and thus enhanced investments in sperm production (Liao et al., 2018; Lüpold et al., 2020). Our results thus reveal how brumation patterns, influenced by environmental variation and physical constraints, affect the socio-ecological context of breeding, the mode and degree of sexual selection, and ultimately the evolution of mating systems, broadening Emlen & Oring’s (1977) general predictions.

In addition to the average size of the testes, their seasonal change also varied with the brumation period. Seasonally breeding anurans regress and regrow their testes between mating seasons (Ogielska and Bartmańska, 2009). Non-brumating species can use energy uptake to compensate for testicular recrudescence, whilst those with a short breeding season after a prolonged inactive period depend on the stored fat to regrow their testes before or immediately after emergence from their hibernaculum. Hence, resources are diverted away from the brain and other organs, especially in species such as Brachytarsophrys spp., in which the fully developed testes combined weigh 12−14 times more than the brain (Data S1).

Brain-testis trade-off

A phylogenetic path analysis confirmed the negative effect of brumation duration on relative brain size and revealed its direct and indirect positive effects on relative testis size, mediated by the amount of adipose tissue, which responded to variation in the inactive period (energetic demand) and the size of the digestive tract (energy uptake). That body fat did not contribute to brain size evolution in this more comprehensive analysis compared to pairwise correlations suggests that the fat−brain trade-off may not be direct. Rather, longer brumation, and thus a short active period, may enhance selection on fat storage for testicular investments in addition to starvation avoidance, while reducing selection for larger brains due to a shifted balance between cognitive benefits and energetic costs (Fig. 3).

A brain-testis trade-off has been reported for bats (Pitnick et al., 2006), but not replicated later in the same (Dechmann and Safi, 2009) or other mammalian taxa (Lemaître et al., 2009). In anurans, the apparent trade-off may result indirectly from opposing selection on brain and testis sizes via environmental seasonality and relative durations of the active and inactive periods. The testes may evolve in response to increased sperm competition and depletion during the shorter and more synchronized breeding season. The brain, while also responding to sexual selection (Mai et al., 2020), is central to various activities, including feeding (Lefebvre et al., 1997) or predator avoidance (Kotrschal et al., 2015; Liao et al., 2022) that are themselves subject to climatic conditions and may independently influence brain evolution. Additionally, whereas testes can regress to save energy when inactive (Ogielska and Bartmańska, 2009), brain metabolism may be less reducible (Mink et al., 1981), resulting in different cost-benefit balances between these organs in relation to seasonality.

Conclusions

In conclusion, our analyses show how brumation in anurans, influenced by high environmental seasonality, may affect resource allocation between costly tissues, directly or through its environmental correlates. The non-independent selective processes promoting diversification in different traits emphasize the need to study the evolutionary trajectory of a given trait such as brain size in the immediate context of both simultaneous investments to other tissues and the species-specific ecology. Our findings also have important implications in the context of sustained environmental change, exacerbated by climate change with its effects on temperature and precipitation patterns around the globe (IPCC, 2022). Such climatic shifts may disrupt the timing of brumation and breeding in many anurans, change their exposure to pathogens or predators through modified activity patterns, affect their ability to find food or suitable breeding sites, or change their population dynamics through resource availability, intra-and interspecific competition, dispersal capabilities, and gene flow (Alves-Ferreira et al., 2022; Blaustein et al., 2010, 2001; Carey and Alexander, 2003). In species with temperature-dependent sexual differentiation, thermal shifts may further change operational sex ratios in breeding populations and thus likely mating dynamics and sexual selection (Eggert, 2004; Lüpold et al., 2017; Ruiz-García et al., 2021). All these factors impose intense environmental pressure on resource acquisition and allocation patterns, and it remains to be seen to what extent variation in the adaptability, and thus resilience, between species exposed to environmental change is attributable to such competing needs between investments and species-specific constraints.

Materials and Methods

Sample collection and preparation

Between 2010 and 2020 and as part of concurrent studies, we collected a total of 396 sexually mature males from 116 anuran species (3.41 ± 0.95 males each) in post-brumation breeding condition and an additional 132 adult males from 50 of these species (2.64 ± 0.94 males each) shortly before entering their hibernacula (Data S1 and S2). For each species, we sampled all males at a single location in southern and western China with known longitude, latitude, and elevation (Data S3). Upon transfer to the laboratory, we sacrificed the individuals by single-pithing, measured their snout-vent length (SVL) to the nearest 0.01 mm with calipers and then preserved them in 4% phosphate-buffered formalin for tissue fixation.

After two months of preservation, we weighed each complete specimen to the nearest 0.1 mg using an electronic balance to obtain body mass before dissecting them following a strict protocol. We separately extracted the brain, heart, liver, lungs, kidneys, spleen, digestive tract, testes, limb muscles, and fat stores, cleaned these tissues and immediately weighed them to the nearest 0.1 mg with an electronic balance. We additionally measured the length of the digestive tract to the nearest 0.01 mm using calipers. We excluded emaciated individuals or those exhibiting visible organ pathologies from our analyses.

Environmental seasonality

For each collection site, we retrieved from the 30-year climate history of https://www.meteoblue.com the monthly mean temperature (in °C) and total precipitation (in mm) (Data S3) and used these values to calculate location-specific annual means and coefficients of variation. We also determined the duration of the dry season, P2T, as the number of months, for which the total precipitation was less than twice the mean temperature (Walter, 1971).

Brumation period

One way that anurans can physiologically respond to seasonality is by adjusting their thermal sensitivity and thus brumation period (Wells, 2007), which in turn could directly or indirectly affect the evolution of brain size (Heldstab et al., 2018). Hence, we estimated the brumation period for all 116 species. To this end, we visited the field sites for 30 of our species daily around the expected start and end times of brumation (based on prior experience). For each species, we recorded the dates and temperatures (using a Kobold HND-T105 high-precision thermometer to the nearest 0.1°C) when the last frogs of a given species were seen at the end of their active period (with no further activity detected for at least seven days) and when the first individuals were detected in the spring. For the same 30 species (and using the same individuals as for morphological measurements), we then experimentally simulated brumation using a Q18 temperature-controlled refrigerator in Shenzhen Pioneer (SAST). We gradually lowered and raised the temperature at a rate of 0.5°C/hour and recorded the temperature at which test subjects entered and left brumation. When entering brumation, anurans drop their heart rate, become sluggish, draw their nictitating membranes across the eyes for protection, spread their legs for stability, and change their physiology to avoid freezing, start breathing through their skin, or switch to anaerobic metabolism (Fei and Ye, 2001; Pinder et al., 1992; Tattersall and Ultsch, 2008; Wells, 2007). To minimize disturbance in our experiment, we used the motion-less four-point stance with the nictitating membranes drawn across the eyes as our proxy of brumation. Our experimental threshold temperatures were tightly associated with the corresponding field measurements both for the start (r = 0.97, t28 = 22.26, P < 0.0001, λ = 0.04 [0.00, 0.47]) and end of the inactive state (r = 0.98, t28 = 28.05, P < 0.0001, λ = 0.04 [0.00, 0.43]). Hence, we assessed the corresponding temperatures for all remaining species in the laboratory and estimated the brumation period based on the daily mean temperatures at the corresponding collection sites as retrieved from Chinese Meteorological Stations (http://www.lishi.tianqi.com) between 2012 and 2016.

We defined the brumation period as the number of consecutive days in each year that remained below this threshold. For simplicity we determined the active rather than brumation period, starting with the first day that the mean daily temperature rose above the activity threshold and remained there for at least five consecutive days, and ending with the last day before the temperature dropped below the activity threshold and remained there until the end of the calendar year. The brumation period then represented the difference between the activity period and the total number of days in each calendar year. Across these five years, the measured temperature thresholds yielded highly repeatable species-specific estimates of the number of days below the activity range (R = 0.95 [95%CI: 0.93−0.96]), as determined by the rpt function in the rptR package (Stoffel et al., 2017) across all 116 species (Fig. S1; Table S1). Further, across the 30 species that were examined both in the lab and the field (see above), these predicted brumation periods were also correlated with the observed brumation periods in the field (r = 0.96, t28 = 18.03, P < 0.0001, λ = 0.05 [0.00, 0.48]; Fig. S8A), which themselves were highly repeatable between years within species (R = 0.98 [0.96−0.99]; Table S1).

Based on this data validation, we used for each species the mean brumation period predicted from our experimentally simulated temperature thresholds. However, to test for potential buffering effects of burrowing in the soil relative to the air temperatures reported by the meteorological stations, we also repeated these estimates by using more conservative thermal thresholds. Here, we restricted the putative brumation days to those with a reported air temperature of either 2°C or 4°C below the experimentally derived inactivity thresholds, simulating prolonged activity by seeking shelter in burrows. The 2°C threshold was based on a pilot study comparing direct measurements of air and burrow temperatures for four different burrows in each of five of our study species (burrow depths: 32.0 ± 3.2 to 121.0 ± 17.8 cm; Fig. S9). Across these species, the burrow-to-air temperature difference reached 1.03 ± 0.35°C to 2.45 ± 0.60°C in measurements around the peak of the brumation period (i.e., early January; Fig. S9). However, since these temporal snapshots were based on sites at relatively low elevation (≤320 m a.s.l.) due to accessibility of burrows during winter, we also used a second, more conservative buffer (4°C below activity range) for comparison. These temperature buffers shortened the predicted brumation periods to a varying degree between species (Fig. S8B); yet the predicted periods covaried strongly between the different temperature thresholds (all r > 0.90, t114 > 21.96, P < 0.0001, all λ < 0.01).

Phylogeny reconstruction

To reconstruct the phylogeny, we obtained the sequences of three nuclear and six mitochondrial genes from GenBank (for accession numbers and sequence coverage see Data S4). The three nuclear genes included the recombination-activating gene 1 (RAG1), rhodopsin (RHOD) and tyrosinase (TYR). The six mitochondrial genes were cytochrome b (CYTB), cytochrome oxidase subunit I (COI), NADH dehydrogenase subunits 2 and 4 (ND2 and ND4), and the large and small subunits of the mitochondrial ribosome genes (12S/16S; omitting the adjacent tRNAs as they were difficult to align and represented only a small amount of data). We aligned the sequences by multi-sequence alignment (MUSCLE) in MEGA v.10.2.2 (Tamura et al., 2013) before comparing possible nucleotide substitution models. The best substitution model, as determined by the function modelTest() in the R (R Core Team, 2022) package phangorn (Schliep, 2011) based on the corrected Akaike Information Criterion, AICc, was GTR+Γ+I for all genes except RHOD, for which HKY+Γ had stronger support.

Using BEAUTi and BEAST v.1.10.4 (Suchard et al., 2018), we then constructed the phylogeny with unlinked substitution models, a relaxed uncorrelated log-normal clock, a Yule speciation process, and the best-supported nucleotide substitution models. We omitted time calibration due to a lack of fossil dates. We ran the Markov Chain Monte Carlo (MCMC) simulation for 55 million generations while sampling every 5,000th tree with a 10% burn-in. Most effective sample size (ESS) values by far exceeded 375 (i.e. all well above the recommended threshold of 200) for all but two tree statistics in the program Tracer v.1.7.2 (Rambaut et al., 2018), thus indicating satisfying convergence of the Bayesian chain and adequate model mixing. Finally, we generated a maximum clade credibility tree with mean node heights and a 10% burn-in using TreeAnnotator v.1.10.4 (Suchard et al., 2018), presented in Fig. S10.

Breeding conditions

To test if a prolonged brumation period reduces the time available for reproduction, thereby changing the level of competition over mates and fertilizations (Lüpold et al., 2017), we extracted the start and end dates of the breeding season from our field notes of concurrent studies on species-specific life histories. These data were available for 43 of our species (Data S3). We used dates when the first and last clutches were observed in focal ponds as a proxy of mating activity, given that males release their sperm during oviposition in these external fertilizers. For each species, dates from at least two years were combined and averaged to obtain the mean duration of the breeding season.

We further recorded whether dense mating aggregations are typically observed in these species. We have previously shown that larger mating clusters, with multiple males clasping the same females, have a significant effect on the evolution of testis size due to the resulting competition among sperm for fertilization (Lüpold et al., 2017). Here, we had no detailed data on the sizes of aggregations and so were only able to code the typical presence or absence of aggregations as a binary variable (Data S3).

Finally, we used our direct estimates of species-specific population densities from our previous study (Lüpold et al., 2017) to test whether a shorter breeding season results in denser breeding populations. Although population density is a more direct measure than the occurrence of aggregations, such data were available for only eight of our species, each based on multiple populations per species (Lüpold et al., 2017). All these data were not necessarily derived from the same years or populations of our main dataset, but given the within-species repeatability in breeding populations (Lüpold et al., 2017) and in the duration of the breeding season (R = 0.88 [0.79−0.93]; Table S1), these differences should be relatively small compared to the interspecific variation and mostly introduce random noise.

Data analyses

General methods

We conducted all statistical analyses in R v.4.2.0 (R Core Team, 2022), using log-transformed data for all phenotypic traits, and for the CV in temperature among the ecological variables. To account for non-independence of data due to common ancestry (Freckleton et al., 2002; Pagel, 1999), we conducted phylogenetic generalized least-squares (PGLS) or phylogenetic logistic regressions (e.g., for occurrence of breeding aggregations), using the R package phylolm (Ho and Ané, 2014) and our reconstructed phylogeny. To account for variation around the species means, we bootstrapped for each model (at 100 fitted replicates) the standardized regression coefficients along with the phylogenetic scaling parameter λ and calculated their corresponding 95% confidence intervals. The λ values indicate phylogenetic independence near zero and strong phylogenetic dependence near one (Freckleton et al., 2002).

Unless stated otherwise, all PGLS models focusing on the relative mass of tissues as the response included snout-vent length (SVL) as a covariate in addition to the focal predictor variable(s). We chose SVL instead of body mass because it is the commonly used measure of body size in anurans and independent of seasonal fluctuations in tissues such as body fat, testes, or limb muscles. One exception, however, was the analysis of phylogenetically informed allometric relationships, for which we cubed SVL such that a slope of 1 equaled unity (isometry). For these allometric relationships we calculated ordinary (generalized) least-squares rather than reduced major-axis regressions, because their greater sensitivity to changes in the steepness, but lower sensitivity to changes in scatter, capture allometric slopes more adequately (Kilmer and Rodríguez, 2017).

Pairwise correlations between tissues

To examine the covariation between different tissues across species, we first calculated pairwise partial correlations controlling for SVL and phylogeny. To this end, we calculated the phylogenetic trait variance-covariance matrix between the pairs of focal variable and SVL using the function phyl.vcv() in phytools (Revell, 2012) with λ = 1 (i.e. Brownian motion), which we then scaled into a correlation matrix using cov2cor() in the stats package (R Core Team, 2022). Using the resulting correlation coefficients rxy, rxz, and ryz, respectively, we then calculated the partial correlation coefficient rxy.z between the x and y variables of interest while accounting for SVL (z) following Crawley’s (2007) equation: rxy.z = with the associated t-statistics and 95% confidence intervals converted using standard conversion and the package effectsize (Ben-Shachar et al., 2020), respectively.

Multivariate allocation patterns

Since pairwise correlations do not necessarily capture more complex, multivariate allocation patterns, we used two additional approaches to explore how tissue sizes varied relative to others: A compositional analysis and a principal component analysis. In both analyses, we focused on those four tissues that covaried with brumation duration or deviated from proportionate scaling with body size: brain, body fat, testes, and hindlimb muscles.

Compositional analyses

In the first approach, we used the function acomp() in the R package compositions (van den Boogaart and Tolosana-Delgado, 2008) to partition total body mass of each species into a five-variable Aitchison composition in a logistic geometry (van den Boogaart and Tolosana-Delgado, 2013), consisting of the proportional representation of the four focal tissues and the remaining body mass combined. Since the focal tissues constituted a size-independent fraction of the total body, the closed composition of this combined mass should be unbiased relative to body size but can instead reveal differential contributions of the four tissues to their total in a multivariate context (Aitchison, 1982; Muldowney et al., 2001; van den Boogaart and Tolosana-Delgado, 2013). For phylogenetic correlations between these variables following the description above, we used centered log ratios obtained by the function clr() in the same package, which maintains the original variable structure. However, owing to the reliance on a full rank of the covariance in multivariate analyses, we used the ilr() function to project the D-part composition isometrically to a D−1 dimensional simplex (Aitchison, 1982), essentially representing the log ratios between the D parts. This multivariate object we subjected to a phylogenetic multivariate regression against brumation duration using the functions mvgls() and manova.gls() in the package mvMORPH (Clavel et al., 2015). For interpretation in the context of the original variable space, we back-transformed the coefficients using the ilrInv() function in compositions.

Phylogenetic principal component analysis

In addition to this compositional data analysis, we also peerformed a phylogenetically informed principal component analysis on the same focal tissues as log-transformed species means, using the phyl.pca() function of the package phytools (Revell, 2012). Here, we primarily focused on the directions of the loading vectors relative to the principal components and one another to glean information on the correlations between the original variables in the principal component space.

Phylogenetic confirmatory path analyses

We performed two phylogenetic confirmatory path analyses (Gonzalez-Voyer and von Hardenberg, 2014; von Hardenberg and Gonzalez-Voyer, 2013) based on pre-specified candidate structural equation models. In the first analysis, we explored the direct and indirect effects of climatic variables on the duration of the breeding season or formation of breeding aggregations. Using eight candidate structural equation models, we essentially tested if breeding aggregations were the result of variation in latitude as a proxy of seasonality, either directly or mediated by the duration of brumation and/or of the breeding season. To avoid overparameterization given that this analysis was based on only 43 species, we did not include testis mass (and so necessarily also SVL to control for body size) as additional variables in the same path models. Rather, to test for links between relative testis size and breeding parameters, we conducted separate directional tests of trait evolution (see below).

A second path analysis aimed to disentangle the different interrelationships between traits that could ultimately mediate the effect of brumation duration on brain and reproductive evolution. Here, we used 28 pre-specified structural equation models to test for direct and indirect links between the brumation period and the four main tissues. As brumation is determined primarily by the environment, we focused on models that explained variation in tissue investments rather than, say, brain size affecting brumation patterns. The effects of brumation on the four tissues were direct or indirect, for example mediated by the digestive tract (resource acquisition) and/or body fat (resource storage), thus providing context for the cognitive buffer, expensive brain, expensive tissue, energy trade-off and fat−brain trade-off hypotheses (Allman et al., 1993; Heldstab et al., 2016; Isler and van Schaik, 2006, 2009; Navarrete et al., 2011). Further, we allowed brain size to affect testis size and vice versa (i.e., expensive sexual tissue hypothesis; Pitnick et al. 2006), and included models in which these organs explained variation in the digestive tract or fat tissue instead of being affected by them (e.g., selection on brain or testis size might mediate selection for greater resource availability rather than resources influencing brain or testis evolution). Across the 28 candidate models, we tested different combinations of these predictions, with traits being explained by single or multiple predictors, or having individual or shared effects on other traits. Using the R package phylopath (van der Bijl, 2018), we examined the conditional independencies of each model, ranked all candidate models based on their C-statistic Information Criterion (CICc), and then averaged the coefficients of the models with ΔCICc ≤ 2 from the top model (von Hardenberg and Gonzalez-Voyer, 2013).

Directional tests of trait evolution

Since the first path analysis did not involve testis size to avoid overparametization, we separately tested for correlated evolution using directional tests of trait evolution (Pagel, 1994; Revell, 2012). One limitation of these models is that they rely on evolutionary transitions between binary states in each trait. Hence, we considered positive residuals of a log-log regression between testis mass and SVL as ‘relatively large testes’ and negative residuals as ‘relatively small testes.’ For the duration of the breeding season, we similarly split the distribution based on the mean duration, whereas aggregation formation was already coded as present or absent. Based on (the weight of) the Akaike Information Criterion (AIC), we then tested if changes in relative testis size and breeding parameters, respectively, were unilaterally dependent, mutually dependent, or independent (Pagel, 1994), using the fitPagel() function in the phytools package (Revell, 2012) with “fitDiscrete” as the optimization method and allowing all rates to differ (i.e., “ARD” model).

Supporting information